---
title: "Piecewise constant hazard model in Stan"
author: "Kazuki Yoshida"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: html_document
---

```{r, message = FALSE, tidy = FALSE, echo = F}
## knitr configuration: http://yihui.name/knitr/options#chunk_options
library(knitr)
showMessage <- FALSE
showWarning <- TRUE
set_alias(w = "fig.width", h = "fig.height", res = "results")
opts_chunk$set(comment = "##", error= TRUE, warning = showWarning, message = showMessage,
               tidy = FALSE, cache = FALSE, echo = TRUE,
               fig.width = 7, fig.height = 7, dev.args = list(family = "sans"))
## for rgl
## knit_hooks$set(rgl = hook_rgl, webgl = hook_webgl)
## for animation
opts_knit$set(animation.fun = hook_ffmpeg_html)
## R configuration
options(width = 116, scipen = 5)
## Record start time
start_time <- Sys.time()
## Configure parallelization
## Parallel backend for foreach (also loads foreach and parallel; includes doMC)
library(doParallel)
## Reproducible parallelization
library(doRNG)
## Detect core count (Do not use on clusters)
n_cores <- parallel::detectCores()
## Used by parallel::mclapply() as default
options(mc.cores = n_cores)
## Used by doParallel as default
options(cores = n_cores)
## Register doParallel as the parallel backend for foreach
## http://stackoverflow.com/questions/28989855/the-difference-between-domc-and-doparallel-in-r
doParallel::registerDoParallel(cores = n_cores)
```

## References
- [(BIDA) Bayesian Ideas and Data Analysis An Introduction for Scientists and Statisticians](http://blogs.oregonstate.edu/bida/) 13.2 Proportional Hazards Modeling
- [(BSA) Bayesian Survival Analysis](https://www.springer.com/us/book/9780387952772) 3.1 Piecewise Constant Hazard Model

## Load packages

```{r}
library(tidyverse)
library(survival)
library(rstan)
library(bayesplot)
set.seed(13960043)
```

### Load and prepare dataset

```
aml                  package:survival                  R Documentation
Acute Myelogenous Leukemia survival data
Description:
     Survival in patients with Acute Myelogenous Leukemia.  The
     question at the time was whether the standard course of
     chemotherapy should be extended ('maintainance') for additional
     cycles.
Usage:
     aml
     leukemia
Format:
       time:    survival or censoring time
       status:  censoring status
       x:       maintenance chemotherapy given? (factor)
Source:
     Rupert G. Miller (1997), _Survival Analysis_.  John Wiley & Sons.
     ISBN: 0-471-25218-2.
```

```{r}
data(leukemia, package = "survival")
leukemia <- as_tibble(leukemia) %>%
    mutate(id = seq_len(n())) %>%
    select(id, everything())
leukemia
```

Check distribution of event times
```{r}
leukemia_summary <- leukemia %>%
    filter(status == 1) %>%
    summarize(n = n(),
              mean_time = mean(time),
              quantiles = list(quantile(time, probs = seq(from = 0, to = 1, by = 0.2)))) %>%
    unnest()
leukemia_summary
```

## Frequentist fit as a references
```{r}
coxph1 <- coxph(formula = Surv(time, status) ~ as.integer(x == "Maintained"),
                data    = leukemia,
                ties    = c("efron","breslow","exact")[1])
summary(coxph1)
```

## Bayesian Model fitting
### Define and compile piecewise constant hazard model
```{r}
## Load and compile
piecewise_ph_model <- rstan::stan_model("./bayesianideas_piecewise2.stan")
piecewise_ph_model
```

### Define time cut points
```{r}
## Cutpoints every 20% of events
cutpoints_20 <- as.numeric(leukemia_summary$quantiles)
## First cutpoint should be time 0.
cutpoints_20[1] <- 0
## Last cutpoint should be larger than the maximum failure time.
cutpoints_20[length(cutpoints_20)] <- cutpoints_20[length(cutpoints_20)] + 1
## Show
cutpoints_20
## Entire time as a single interval for exponential model
cutpoints_100 <- c(0, max(cutpoints_20))
cutpoints_100
## All unique event times
cutpoints_all <- unique(sort(leukemia$time[leukemia$status == 1]))
cutpoints_all <- c(0, cutpoints_all, max(cutpoints_all)+1)
cutpoints_all
## Evaluation grid for survival function
grid <- seq(from = 0, to = max(leukemia_summary$quantiles), by = 0.1)
```

### Exponential model (Prior)
```{r,results="hide"}
exponential_prior_sample <-
    rstan::sampling(object = piecewise_ph_model,
                    data = list(w1 = 10^4,
                                lambda1_star = 0.01,
                                w = 0.01,
                                lambda_star = 0.05,
                                beta_mean = 0,
                                beta_sd = 10,
                                K = length(cutpoints_100) - 1,
                                cutpoints = cutpoints_100,
                                N = length(leukemia$time),
                                cens = leukemia$status,
                                y = leukemia$time,
                                x = as.integer(leukemia$x == "Maintained"),
                                grid_size = length(grid),
                                grid = grid,
                                ## Do not use likelihood
                                eval_likelihood = 0))
```
```{r}
check_hmc_diagnostics(exponential_prior_sample)
print(exponential_prior_sample, pars = c("lambda","beta","lp__"))
```
```{r}
traceplot(exponential_prior_sample, inc_warmup = TRUE, pars = c("lambda","beta","lp__"))
traceplot(exponential_prior_sample, inc_warmup = FALSE, pars = c("lambda","beta","lp__"))
bayesplot::mcmc_rank_overlay(exponential_prior_sample, regex_pars = c("lambda","beta","lp__"))
```
Baseline hazard function.
```{r}
exponential_prior_sample %>%
    as.matrix(pars = "h0_grid") %>%
    as_tibble() %>%
    `names<-`(as.character(grid)) %>%
    mutate(iter = seq_len(n())) %>%
    gather(key = time, value = value, -iter) %>%
    mutate(time = as.numeric(time)) %>%
    filter(iter %in% sample(1:max(iter), size = 500)) %>%
    ##
    ggplot(mapping = aes(x = time, y = value, group = iter)) +
    geom_line(alpha = 0.1) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
          legend.key = element_blank(),
          plot.title = element_text(hjust = 0.5),
          strip.background = element_blank())
```
Baseline cumulative hazard function.
```{r}
exponential_prior_sample %>%
    as.matrix(pars = "H0_grid") %>%
    as_tibble() %>%
    `names<-`(as.character(grid)) %>%
    mutate(iter = seq_len(n())) %>%
    gather(key = time, value = value, -iter) %>%
    mutate(time = as.numeric(time)) %>%
    filter(iter %in% sample(1:max(iter), size = 500)) %>%
    ##
    ggplot(mapping = aes(x = time, y = value, group = iter)) +
    geom_line(alpha = 0.1) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
          legend.key = element_blank(),
          plot.title = element_text(hjust = 0.5),
          strip.background = element_blank())
```
Baseline survival estimate for the Nonmaintained group.
```{r}
exponential_prior_sample %>%
    as.matrix(pars = "S0_grid") %>%
    as_tibble() %>%
    `names<-`(as.character(grid)) %>%
    mutate(iter = seq_len(n())) %>%
    gather(key = time, value = value, -iter) %>%
    mutate(time = as.numeric(time)) %>%
    filter(iter %in% sample(1:max(iter), size = 500)) %>%
    ##
    ggplot(mapping = aes(x = time, y = value, group = iter)) +
    geom_line(alpha = 0.1) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
          legend.key = element_blank(),
          plot.title = element_text(hjust = 0.5),
          strip.background = element_blank())
```
Survival estimate for the Maintained group.
```{r}
exponential_prior_sample %>%
    as.matrix(pars = "S1_grid") %>%
    as_tibble() %>%
    `names<-`(as.character(grid)) %>%
    mutate(iter = seq_len(n())) %>%
    gather(key = time, value = value, -iter) %>%
    mutate(time = as.numeric(time)) %>%
    filter(iter %in% sample(1:max(iter), size = 500)) %>%
    ##
    ggplot(mapping = aes(x = time, y = value, group = iter)) +
    geom_line(alpha = 0.1) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
          legend.key = element_blank(),
          plot.title = element_text(hjust = 0.5),
          strip.background = element_blank())
```

### Exponential model (Posterior)
```{r,results="hide"}
exponential_sample <-
    rstan::sampling(object = piecewise_ph_model,
                    data = list(w1 = 10^4,
                                lambda1_star = 0.01,
                                w = 0.01,
                                lambda_star = 0.05,
                                beta_mean = 0,
                                beta_sd = 10,
                                K = length(cutpoints_100) - 1,
                                cutpoints = cutpoints_100,
                                N = length(leukemia$time),
                                cens = leukemia$status,
                                y = leukemia$time,
                                x = as.integer(leukemia$x == "Maintained"),
                                grid_size = length(grid),
                                grid = grid,
                                ## Do use likelihood
                                eval_likelihood = 1))
```
```{r}
check_hmc_diagnostics(exponential_sample)
print(exponential_sample, pars = c("lambda","beta","lp__"))
```
```{r}
traceplot(exponential_sample, inc_warmup = TRUE, pars = c("lambda","beta","lp__"))
traceplot(exponential_sample, inc_warmup = FALSE, pars = c("lambda","beta","lp__"))
bayesplot::mcmc_rank_overlay(exponential_sample, regex_pars = c("lambda","beta","lp__"))
```
Baseline hazard function.
```{r}
exponential_sample %>%
    as.matrix(pars = "h0_grid") %>%
    as_tibble() %>%
    `names<-`(as.character(grid)) %>%
    mutate(iter = seq_len(n())) %>%
    gather(key = time, value = value, -iter) %>%
    mutate(time = as.numeric(time)) %>%
    filter(iter %in% sample(1:max(iter), size = 500)) %>%
    ##
    ggplot(mapping = aes(x = time, y = value, group = iter)) +
    geom_line(alpha = 0.1) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
          legend.key = element_blank(),
          plot.title = element_text(hjust = 0.5),
          strip.background = element_blank())
```
Baseline cumulative hazard function.
```{r}
exponential_sample %>%
    as.matrix(pars = "H0_grid") %>%
    as_tibble() %>%
    `names<-`(as.character(grid)) %>%
    mutate(iter = seq_len(n())) %>%
    gather(key = time, value = value, -iter) %>%
    mutate(time = as.numeric(time)) %>%
    filter(iter %in% sample(1:max(iter), size = 500)) %>%
    ##
    ggplot(mapping = aes(x = time, y = value, group = iter)) +
    geom_line(alpha = 0.1) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
          legend.key = element_blank(),
          plot.title = element_text(hjust = 0.5),
          strip.background = element_blank())
```
Baseline survival estimate for the Nonmaintained group.
```{r}
exponential_sample %>%
    as.matrix(pars = "S0_grid") %>%
    as_tibble() %>%
    `names<-`(as.character(grid)) %>%
    mutate(iter = seq_len(n())) %>%
    gather(key = time, value = value, -iter) %>%
    mutate(time = as.numeric(time)) %>%
    filter(iter %in% sample(1:max(iter), size = 500)) %>%
    ##
    ggplot(mapping = aes(x = time, y = value, group = iter)) +
    geom_line(alpha = 0.1) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
          legend.key = element_blank(),
          plot.title = element_text(hjust = 0.5),
          strip.background = element_blank())
```
Survival estimate for the Maintained group.
```{r}
exponential_sample %>%
    as.matrix(pars = "S1_grid") %>%
    as_tibble() %>%
    `names<-`(as.character(grid)) %>%
    mutate(iter = seq_len(n())) %>%
    gather(key = time, value = value, -iter) %>%
    mutate(time = as.numeric(time)) %>%
    filter(iter %in% sample(1:max(iter), size = 500)) %>%
    ##
    ggplot(mapping = aes(x = time, y = value, group = iter)) +
    geom_line(alpha = 0.1) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
          legend.key = element_blank(),
          plot.title = element_text(hjust = 0.5),
          strip.background = element_blank())
```

### Time cutoffs at 20% quantiles (Prior)
```{r,results="hide"}
piecewise_ph_prior_sample <-
    rstan::sampling(object = piecewise_ph_model,
                    data = list(w1 = 10^4,
                                lambda1_star = 0.01,
                                w = 0.01,
                                lambda_star = 0.05,
                                beta_mean = 0,
                                beta_sd = 10,
                                K = length(cutpoints_20) - 1,
                                cutpoints = cutpoints_20,
                                N = length(leukemia$time),
                                cens = leukemia$status,
                                y = leukemia$time,
                                x = as.integer(leukemia$x == "Maintained"),
                                grid_size = length(grid),
                                grid = grid,
                                ## Do not use likelihood
                                eval_likelihood = 0))
```
```{r}
check_hmc_diagnostics(piecewise_ph_prior_sample)
print(piecewise_ph_prior_sample, pars = c("lambda","beta","lp__"))
```
```{r}
traceplot(piecewise_ph_prior_sample, inc_warmup = TRUE, pars = c("lambda","beta","lp__"))
traceplot(piecewise_ph_prior_sample, inc_warmup = FALSE, pars = c("lambda","beta","lp__"))
bayesplot::mcmc_rank_overlay(piecewise_ph_prior_sample, regex_pars = c("lambda","beta","lp__"))
```
Baseline hazard function.
```{r}
piecewise_ph_prior_sample %>%
    as.matrix(pars = "h0_grid") %>%
    as_tibble() %>%
    `names<-`(as.character(grid)) %>%
    mutate(iter = seq_len(n())) %>%
    gather(key = time, value = value, -iter) %>%
    mutate(time = as.numeric(time)) %>%
    filter(iter %in% sample(1:max(iter), size = 500)) %>%
    ##
    ggplot(mapping = aes(x = time, y = value, group = iter)) +
    geom_line(alpha = 0.1) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
          legend.key = element_blank(),
          plot.title = element_text(hjust = 0.5),
          strip.background = element_blank())
```
Baseline cumulative hazard function.
```{r}
piecewise_ph_prior_sample %>%
    as.matrix(pars = "H0_grid") %>%
    as_tibble() %>%
    `names<-`(as.character(grid)) %>%
    mutate(iter = seq_len(n())) %>%
    gather(key = time, value = value, -iter) %>%
    mutate(time = as.numeric(time)) %>%
    filter(iter %in% sample(1:max(iter), size = 500)) %>%
    ##
    ggplot(mapping = aes(x = time, y = value, group = iter)) +
    geom_line(alpha = 0.1) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
          legend.key = element_blank(),
          plot.title = element_text(hjust = 0.5),
          strip.background = element_blank())
```
Baseline survival estimate for the Nonmaintained group.
```{r}
piecewise_ph_prior_sample %>%
    as.matrix(pars = "S0_grid") %>%
    as_tibble() %>%
    `names<-`(as.character(grid)) %>%
    mutate(iter = seq_len(n())) %>%
    gather(key = time, value = value, -iter) %>%
    mutate(time = as.numeric(time)) %>%
    filter(iter %in% sample(1:max(iter), size = 500)) %>%
    ##
    ggplot(mapping = aes(x = time, y = value, group = iter)) +
    geom_line(alpha = 0.1) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
          legend.key = element_blank(),
          plot.title = element_text(hjust = 0.5),
          strip.background = element_blank())
```
Survival estimate for the Maintained group.
```{r}
piecewise_ph_prior_sample %>%
    as.matrix(pars = "S1_grid") %>%
    as_tibble() %>%
    `names<-`(as.character(grid)) %>%
    mutate(iter = seq_len(n())) %>%
    gather(key = time, value = value, -iter) %>%
    mutate(time = as.numeric(time)) %>%
    filter(iter %in% sample(1:max(iter), size = 500)) %>%
    ##
    ggplot(mapping = aes(x = time, y = value, group = iter)) +
    geom_line(alpha = 0.1) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
          legend.key = element_blank(),
          plot.title = element_text(hjust = 0.5),
          strip.background = element_blank())
```

### Time cutoffs at 20% quantiles (Posterior)
```{r,results="hide"}
piecewise_ph_sample <-
    rstan::sampling(object = piecewise_ph_model,
                    data = list(w1 = 10^4,
                                lambda1_star = 0.01,
                                w = 0.01,
                                lambda_star = 0.05,
                                beta_mean = 0,
                                beta_sd = 10,
                                K = length(cutpoints_20) - 1,
                                cutpoints = cutpoints_20,
                                N = length(leukemia$time),
                                cens = leukemia$status,
                                y = leukemia$time,
                                x = as.integer(leukemia$x == "Maintained"),
                                grid_size = length(grid),
                                grid = grid,
                                ## Do use likelihood
                                eval_likelihood = 1))
```
```{r}
check_hmc_diagnostics(piecewise_ph_sample)
print(piecewise_ph_sample, pars = c("lambda","beta","lp__"))
```
```{r}
traceplot(piecewise_ph_sample, inc_warmup = TRUE, pars = c("lambda","beta","lp__"))
traceplot(piecewise_ph_sample, inc_warmup = FALSE, pars = c("lambda","beta","lp__"))
bayesplot::mcmc_rank_overlay(piecewise_ph_sample, regex_pars = c("lambda","beta","lp__"))
```
Baseline hazard function.
```{r}
piecewise_ph_sample %>%
    as.matrix(pars = "h0_grid") %>%
    as_tibble() %>%
    `names<-`(as.character(grid)) %>%
    mutate(iter = seq_len(n())) %>%
    gather(key = time, value = value, -iter) %>%
    mutate(time = as.numeric(time)) %>%
    filter(iter %in% sample(1:max(iter), size = 500)) %>%
    ##
    ggplot(mapping = aes(x = time, y = value, group = iter)) +
    geom_line(alpha = 0.1) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
          legend.key = element_blank(),
          plot.title = element_text(hjust = 0.5),
          strip.background = element_blank())
```
Baseline cumulative hazard function.
```{r}
piecewise_ph_sample %>%
    as.matrix(pars = "H0_grid") %>%
    as_tibble() %>%
    `names<-`(as.character(grid)) %>%
    mutate(iter = seq_len(n())) %>%
    gather(key = time, value = value, -iter) %>%
    mutate(time = as.numeric(time)) %>%
    filter(iter %in% sample(1:max(iter), size = 500)) %>%
    ##
    ggplot(mapping = aes(x = time, y = value, group = iter)) +
    geom_line(alpha = 0.1) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
          legend.key = element_blank(),
          plot.title = element_text(hjust = 0.5),
          strip.background = element_blank())
```
Baseline survival estimate for the Nonmaintained group.
```{r}
piecewise_ph_sample %>%
    as.matrix(pars = "S0_grid") %>%
    as_tibble() %>%
    `names<-`(as.character(grid)) %>%
    mutate(iter = seq_len(n())) %>%
    gather(key = time, value = value, -iter) %>%
    mutate(time = as.numeric(time)) %>%
    filter(iter %in% sample(1:max(iter), size = 500)) %>%
    ##
    ggplot(mapping = aes(x = time, y = value, group = iter)) +
    geom_line(alpha = 0.1) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
          legend.key = element_blank(),
          plot.title = element_text(hjust = 0.5),
          strip.background = element_blank())
```
Survival estimate for the Maintained group.
```{r}
piecewise_ph_sample %>%
    as.matrix(pars = "S1_grid") %>%
    as_tibble() %>%
    `names<-`(as.character(grid)) %>%
    mutate(iter = seq_len(n())) %>%
    gather(key = time, value = value, -iter) %>%
    mutate(time = as.numeric(time)) %>%
    filter(iter %in% sample(1:max(iter), size = 500)) %>%
    ##
    ggplot(mapping = aes(x = time, y = value, group = iter)) +
    geom_line(alpha = 0.1) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
          legend.key = element_blank(),
          plot.title = element_text(hjust = 0.5),
          strip.background = element_blank())
```

### Time cutoffs at all event times (Prior)
```{r,results="hide"}
piecewise_ph_all_prior_sample <-
    rstan::sampling(object = piecewise_ph_model,
                    data = list(w1 = 10^4,
                                lambda1_star = 0.01,
                                w = 0.01,
                                lambda_star = 0.05,
                                beta_mean = 0,
                                beta_sd = 10,
                                K = length(cutpoints_all) - 1,
                                cutpoints = cutpoints_all,
                                N = length(leukemia$time),
                                cens = leukemia$status,
                                y = leukemia$time,
                                x = as.integer(leukemia$x == "Maintained"),
                                grid_size = length(grid),
                                grid = grid,
                                eval_likelihood = 1))
```
```{r}
check_hmc_diagnostics(piecewise_ph_all_prior_sample)
print(piecewise_ph_all_prior_sample, pars = c("lambda","beta","lp__"))
```
```{r}
traceplot(piecewise_ph_all_prior_sample, inc_warmup = TRUE, pars = c("lambda","beta","lp__"))
traceplot(piecewise_ph_all_prior_sample, inc_warmup = FALSE, pars = c("lambda","beta","lp__"))
bayesplot::mcmc_rank_overlay(piecewise_ph_all_prior_sample, regex_pars = c("lambda","beta","lp__"))
```
Baseline hazard function.
```{r}
piecewise_ph_all_prior_sample %>%
    as.matrix(pars = "h0_grid") %>%
    as_tibble() %>%
    `names<-`(as.character(grid)) %>%
    mutate(iter = seq_len(n())) %>%
    gather(key = time, value = value, -iter) %>%
    mutate(time = as.numeric(time)) %>%
    filter(iter %in% sample(1:max(iter), size = 500)) %>%
    ##
    ggplot(mapping = aes(x = time, y = value, group = iter)) +
    geom_line(alpha = 0.1) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
          legend.key = element_blank(),
          plot.title = element_text(hjust = 0.5),
          strip.background = element_blank())
```
Baseline cumulative hazard function.
```{r}
piecewise_ph_all_prior_sample %>%
    as.matrix(pars = "H0_grid") %>%
    as_tibble() %>%
    `names<-`(as.character(grid)) %>%
    mutate(iter = seq_len(n())) %>%
    gather(key = time, value = value, -iter) %>%
    mutate(time = as.numeric(time)) %>%
    filter(iter %in% sample(1:max(iter), size = 500)) %>%
    ##
    ggplot(mapping = aes(x = time, y = value, group = iter)) +
    geom_line(alpha = 0.1) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
          legend.key = element_blank(),
          plot.title = element_text(hjust = 0.5),
          strip.background = element_blank())
```
Baseline survival estimate for the Nonmaintained group.
```{r}
piecewise_ph_all_prior_sample %>%
    as.matrix(pars = "S0_grid") %>%
    as_tibble() %>%
    `names<-`(as.character(grid)) %>%
    mutate(iter = seq_len(n())) %>%
    gather(key = time, value = value, -iter) %>%
    mutate(time = as.numeric(time)) %>%
    filter(iter %in% sample(1:max(iter), size = 500)) %>%
    ##
    ggplot(mapping = aes(x = time, y = value, group = iter)) +
    geom_line(alpha = 0.1) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
          legend.key = element_blank(),
          plot.title = element_text(hjust = 0.5),
          strip.background = element_blank())
```
Survival estimate for the Maintained group.
```{r}
piecewise_ph_all_prior_sample %>%
    as.matrix(pars = "S1_grid") %>%
    as_tibble() %>%
    `names<-`(as.character(grid)) %>%
    mutate(iter = seq_len(n())) %>%
    gather(key = time, value = value, -iter) %>%
    mutate(time = as.numeric(time)) %>%
    filter(iter %in% sample(1:max(iter), size = 500)) %>%
    ##
    ggplot(mapping = aes(x = time, y = value, group = iter)) +
    geom_line(alpha = 0.1) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
          legend.key = element_blank(),
          plot.title = element_text(hjust = 0.5),
          strip.background = element_blank())
```

### Time cutoffs at all event times (Posterior)
```{r,results="hide"}
piecewise_ph_all_sample <-
    rstan::sampling(object = piecewise_ph_model,
                    data = list(w1 = 10^4,
                                lambda1_star = 0.01,
                                w = 0.01,
                                lambda_star = 0.05,
                                beta_mean = 0,
                                beta_sd = 10,
                                K = length(cutpoints_all) - 1,
                                cutpoints = cutpoints_all,
                                N = length(leukemia$time),
                                cens = leukemia$status,
                                y = leukemia$time,
                                x = as.integer(leukemia$x == "Maintained"),
                                grid_size = length(grid),
                                grid = grid,
                                eval_likelihood = 1))
```
```{r}
check_hmc_diagnostics(piecewise_ph_all_sample)
print(piecewise_ph_all_sample, pars = c("lambda","beta","lp__"))
```
```{r}
traceplot(piecewise_ph_all_sample, inc_warmup = TRUE, pars = c("lambda","beta","lp__"))
traceplot(piecewise_ph_all_sample, inc_warmup = FALSE, pars = c("lambda","beta","lp__"))
bayesplot::mcmc_rank_overlay(piecewise_ph_all_sample, regex_pars = c("lambda","beta","lp__"))
```
Baseline hazard function.
```{r}
piecewise_ph_all_sample %>%
    as.matrix(pars = "h0_grid") %>%
    as_tibble() %>%
    `names<-`(as.character(grid)) %>%
    mutate(iter = seq_len(n())) %>%
    gather(key = time, value = value, -iter) %>%
    mutate(time = as.numeric(time)) %>%
    filter(iter %in% sample(1:max(iter), size = 500)) %>%
    ##
    ggplot(mapping = aes(x = time, y = value, group = iter)) +
    geom_line(alpha = 0.1) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
          legend.key = element_blank(),
          plot.title = element_text(hjust = 0.5),
          strip.background = element_blank())
```
Baseline cumulative hazard function.
```{r}
piecewise_ph_all_sample %>%
    as.matrix(pars = "H0_grid") %>%
    as_tibble() %>%
    `names<-`(as.character(grid)) %>%
    mutate(iter = seq_len(n())) %>%
    gather(key = time, value = value, -iter) %>%
    mutate(time = as.numeric(time)) %>%
    filter(iter %in% sample(1:max(iter), size = 500)) %>%
    ##
    ggplot(mapping = aes(x = time, y = value, group = iter)) +
    geom_line(alpha = 0.1) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
          legend.key = element_blank(),
          plot.title = element_text(hjust = 0.5),
          strip.background = element_blank())
```
Baseline survival estimate for the Nonmaintained group.
```{r}
piecewise_ph_all_sample %>%
    as.matrix(pars = "S0_grid") %>%
    as_tibble() %>%
    `names<-`(as.character(grid)) %>%
    mutate(iter = seq_len(n())) %>%
    gather(key = time, value = value, -iter) %>%
    mutate(time = as.numeric(time)) %>%
    filter(iter %in% sample(1:max(iter), size = 500)) %>%
    ##
    ggplot(mapping = aes(x = time, y = value, group = iter)) +
    geom_line(alpha = 0.1) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
          legend.key = element_blank(),
          plot.title = element_text(hjust = 0.5),
          strip.background = element_blank())
```
Survival estimate for the Maintained group.
```{r}
piecewise_ph_all_sample %>%
    as.matrix(pars = "S1_grid") %>%
    as_tibble() %>%
    `names<-`(as.character(grid)) %>%
    mutate(iter = seq_len(n())) %>%
    gather(key = time, value = value, -iter) %>%
    mutate(time = as.numeric(time)) %>%
    filter(iter %in% sample(1:max(iter), size = 500)) %>%
    ##
    ggplot(mapping = aes(x = time, y = value, group = iter)) +
    geom_line(alpha = 0.1) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
          legend.key = element_blank(),
          plot.title = element_text(hjust = 0.5),
          strip.background = element_blank())
```

--------------------
- Top Page: http://rpubs.com/kaz_yos/
- Github: https://github.com/kaz-yos

```{r}
print(sessionInfo())
## Record execution time and multicore use
end_time <- Sys.time()
diff_time <- difftime(end_time, start_time, units = "auto")
cat("Started  ", as.character(start_time), "\n",
    "Finished ", as.character(end_time), "\n",
    "Time difference of ", diff_time, " ", attr(diff_time, "units"), "\n",
    "Used ", foreach::getDoParWorkers(), " cores\n",
    "Used ", foreach::getDoParName(), " as backend\n",
    sep = "")
```
