---
title: "BIDA: Density Estimation (Finite Mixture Model)"
author: "Kazuki Yoshida"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: html_document
---

```{r, message = FALSE, tidy = FALSE, echo = F}
## knitr configuration: http://yihui.name/knitr/options#chunk_options
library(knitr)
showMessage <- FALSE
showWarning <- TRUE
set_alias(w = "fig.width", h = "fig.height", res = "results")
opts_chunk$set(comment = "##", error= TRUE, warning = showWarning, message = showMessage,
               tidy = FALSE, cache = TRUE, echo = TRUE,
               fig.width = 7, fig.height = 7, dev.args = list(family = "sans"))
## for rgl
## knit_hooks$set(rgl = hook_rgl, webgl = hook_webgl)
## for animation
opts_knit$set(animation.fun = hook_ffmpeg_html)
## R configuration
options(width = 116, scipen = 5)
## Record start time
start_time <- Sys.time()
## Configure parallelization
## Parallel backend for foreach (also loads foreach and parallel; includes doMC)
library(doParallel)
## Reproducible parallelization
library(doRNG)
## Detect core count (Do not use on clusters)
n_cores <- parallel::detectCores()
## Used by parallel::mclapply() as default
options(mc.cores = n_cores)
## Used by doParallel as default
options(cores = n_cores)
## Register doParallel as the parallel backend for foreach
## http://stackoverflow.com/questions/28989855/the-difference-between-domc-and-doparallel-in-r
doParallel::registerDoParallel(cores = n_cores)
```

## References
- Books
  - [(BIDA) Bayesian Ideas and Data Analysis An Introduction for Scientists and Statisticians](http://blogs.oregonstate.edu/bida/)
  - [(BDA) Bayesian Data Analysis](http://www.stat.columbia.edu/~gelman/book/)
- Web sites
  - [Bayesian Statistics Using Stan 10 Finite Mixtures](http://www.stat.columbia.edu/~gelman/bda.course/_book/mixture-modeling-chapter.html)
  - [Finite mixture models in Stan](http://modernstatisticalworkflow.blogspot.com/2016/10/finite-mixture-models-in-stan.html)
  - [Identifying Bayesian Mixture Models](http://mc-stan.org/users/documentation/case-studies/identifying_mixture_models.html)
  - [Taming Divergences in Stan Models](https://www.martinmodrak.cz/2018/02/19/taming-divergences-in-stan-models/)
  - [Mixture models in Stan: you can use log_mix](https://andrewgelman.com/2017/08/21/mixture-models-stan-can-use-log_mix/)
  - [Identification of mixture of multivariate normal distributions](https://discourse.mc-stan.org/t/identification-of-mixture-of-multivariate-normal-distributions/4203)
- jSoftware
  - [Stan Modeling Language User's Guide and Reference Manual, Version 2.17.0](http://mc-stan.org/users/documentation/)
  - [CRAN DPpackage: Bayesian Nonparametric Modeling in R](https://cran.r-project.org/package=DPpackage)
  - [J Stat Softw. 2011. DPpackage: Bayesian Non- and Semi-parametric Modelling in R.](https://www.jstatsoft.org/article/view/v040i05)


## Load packages
```{r}
library(tidyverse)
library(rstan)
## devtools::install_github('jburos/biostan', build_vignettes = TRUE, dependencies = TRUE)
## library(biostan)
set.seed(732565397)
```

## Prepare data

We generate a dataset with clearly separate modes.

```{r}
data1 <- data_frame(z = 1 + rbinom(n = 82, size = 1, prob = 0.3),
           x1 = rnorm(n = length(z), mean = -10, sd = 1),
           x2 = rnorm(n = length(z), mean = +8, sd = 2),
           x = ifelse(z == 1, x1, x2))
ggplot(data = data1, mapping = aes(x = x)) +
    geom_point(y = 0.5) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
          legend.key = element_blank(),
          plot.title = element_text(hjust = 0.5),
          strip.background = element_blank())
```

Set range of values to examine estimated density.

```{r}
grid_max <- 20
grid_min <- -20
grid_length <- 100
```

Define some helper functions

```{r}
plot_draws <- function(stan_fit) {
    ## Note direct access to global variables
    draw_data  <- tidybayes::tidy_draws(stan_fit) %>%
        select(.chain, .iteration, .draw, starts_with("log_f")) %>%
        gather(key = key, value = value, starts_with("log_f")) %>%
        mutate(key = gsub("log_f|\\[|\\]", "", key) %>% as.integer(),
           x = factor(key, labels = seq(from = grid_min, to = grid_max, length.out = grid_length)) %>%
               as.character() %>%
               as.numeric(),
           value = exp(value))

    summary_density <- draw_data %>%
        group_by(.chain, x) %>%
        summarize(value = mean(value))

    ggplot(data = draw_data, mapping = aes(x = x, y = value,
           group = interaction(.chain, .iteration, .draw))) +
    geom_line(size = 0.1, alpha = 1/20) +
    geom_line(data = summary_density, mapping = aes(group = .chain), size = 0.5, color = "gray") +
    geom_point(data = galaxy, mapping = aes(x = k_speed, group = NULL), y = 0) +
    facet_wrap(~ .chain) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
          legend.key = element_blank(),
          plot.title = element_text(hjust = 0.5),
          strip.background = element_blank())
}

traceplot_all <- function(fit) {
    print(traceplot(fit, inc_warmup = TRUE, pars = "mu"))
    print(traceplot(fit, inc_warmup = TRUE, pars = "sigma_squared"))
    print(traceplot(fit, inc_warmup = TRUE, pars = "Pi"))
    print(traceplot(fit, inc_warmup = FALSE, pars = "lp__"))
}

pairs_plot_all <- function(fit) {
    pairs(fit, pars = "mu")
    pairs(fit, pars = "sigma_squared")
    pairs(fit, pars = "Pi")
}

```


## Finite Mixture of Normals

### Model
Let $z_{i} \in \left\{ 1, \dots, H \right\}$ be the cluster membership latent variable and $\mathbf{z}_{i} = (I(z_{i}=1),\dots,I(z_{i}=H))^{T}$ be the vector indicator version. We have the following model.

$$\begin{align*}
  y_{i} | (z_{i}, \mu_{1},\dots,\mu_{H}, \sigma^{2}_{1},\dots,\sigma^{2}_{H}) &\sim N(\mu_{z_{i}}, \sigma^{2}_{z_{i}})\\
  \mathbf{z}_{i} | \boldsymbol{\pi} &\sim Multinomial(1, \boldsymbol{\pi})\\
  \\
  p(\mu_{h},\sigma^{2}_{h}) &= p(\mu_{h}|\sigma^{2}_{h})p(\sigma^{2}_{h})\\
  &= N(\mu_{h} | m, s^{2}) Inverse-Gamma(\sigma^{2}_{h} | \alpha, \beta)\\
  &= \left[ \frac{1}{\sqrt{2\pi s^{2}}} \exp \left( - \frac{(\mu_{h} - m)^{2}}{2 \times s^{2}} \right) \right]
    \left[ \frac{\beta^{\alpha}}{\Gamma(\alpha)} (\sigma^{2}_{h})^{-\alpha - 1} \exp \left( - \frac{\beta}{\sigma^{2}_{h}} \right) \right]\\
  &\text{where}\\
  m &= 0\\
  s^{2} &= 10^{3}\\
  \alpha &= 10^{-3}\\
  \beta &= 10^{-3}\\
  \\
  p(\boldsymbol{\pi}) &\sim Dirichlet \left( \frac{\alpha}{H}, \dots, \frac{\alpha}{H} \right)\\
  \end{align*}
$$

Summing out the latent categorical variable $z_{i}$ results in the following (conditioning on parameters suppressed for simplicity) marginal density. Note the cluster membership latent variable $z_i$ is not measured, thus, it is a discrete parameter. Stan's Hamiltonian Monte Carlo (HMC) cannot deal with discrete parameters, this marginalization step is required fro Stan. JAGS seems to allow a discrete parameter and accepts the original model above.

$$\begin{align*}
  p(y) &= \sum^{H}_{z=1} p(y|z) p(z)\\
  &= \sum^{H}_{z=1} \pi_{z} p(y|z)\\
  &= \sum^{H}_{h=1} \pi_h N(y | \mu_{h}, \sigma^{2}_{h})\\
  \end{align*}
$$


### Fitting
#### 2 latent cluster model
```{r}
normal_fixed_mixture_unordered_stan_code <- biostan::read_stan_file("./bayesianideas_density_normal_fixed_mixture_unordered.stan")
biostan::print_stan_code(normal_fixed_mixture_unordered_stan_code, section = NULL)
```
```{r, cache = TRUE, results = 'hide'}
normal_fixed_mixture_unordered_stan_fit <-
    rstan::stan(model_code = normal_fixed_mixture_unordered_stan_code,
                data = list(alpha = 10^(-3), beta = 10^(-3),
                            m = 0, s_squared = 10^(3),
                            n = nrow(data1),
                            y = data1$x,
                            dirichlet_alpha = 1,
                            H = 2,
                            grid_max = grid_max, grid_min = grid_min, grid_length = grid_length))
```
```{r}
normal_fixed_mixture_unordered_stan_fit
traceplot_all(normal_fixed_mixture_unordered_stan_fit)
plot_draws(normal_fixed_mixture_unordered_stan_fit)
pairs_plot_all(normal_fixed_mixture_unordered_stan_fit)
```


--------------------
- Top Page: http://rpubs.com/kaz_yos/
- Github: https://github.com/kaz-yos

```{r}
print(sessionInfo())
## Record execution time and multicore use
end_time <- Sys.time()
diff_time <- difftime(end_time, start_time, units = "auto")
cat("Started  ", as.character(start_time), "\n",
    "Finished ", as.character(end_time), "\n",
    "Time difference of ", diff_time, " ", attr(diff_time, "units"), "\n",
    "Used ", foreach::getDoParWorkers(), " cores\n",
    "Used ", foreach::getDoParName(), " as backend\n",
    sep = "")
```
