---
title: "Bayesian Causal Inference"
author: "Kazuki Yoshida"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: html_document
---

```{r, message = FALSE, tidy = FALSE, echo = F}
## knitr configuration: http://yihui.name/knitr/options#chunk_options
library(knitr)
showMessage <- FALSE
showWarning <- TRUE
set_alias(w = "fig.width", h = "fig.height", res = "results")
opts_chunk$set(comment = "##", error= TRUE, warning = showWarning, message = showMessage,
               tidy = FALSE, cache = FALSE, echo = TRUE,
               fig.width = 7, fig.height = 7, dev.args = list(family = "sans"))
## for rgl
## knit_hooks$set(rgl = hook_rgl, webgl = hook_webgl)
## for animation
opts_knit$set(animation.fun = hook_ffmpeg_html)
## R configuration
options(width = 116, scipen = 5)
## Record start time
start_time <- Sys.time()
## Configure parallelization
## Parallel backend for foreach (also loads foreach and parallel; includes doMC)
library(doParallel)
## Reproducible parallelization
library(doRNG)
## Detect core count (Do not use on clusters)
n_cores <- parallel::detectCores()
## Used by parallel::mclapply() as default
options(mc.cores = n_cores)
## Used by doParallel as default
options(cores = n_cores)
## Register doParallel as the parallel backend for foreach
## http://stackoverflow.com/questions/28989855/the-difference-between-domc-and-doparallel-in-r
doParallel::registerDoParallel(cores = n_cores)
```

## References
-

## Load packages
Here we use the ```distributed``` that I wrote for a previous simulation project ([Comparison of Privacy-Protecting Analytic and Data-sharing Methods: a Simulation Study](https://github.com/kaz-yos/distributed)).
```{r}
library(tidyverse)
## Simulation suite from
## https://github.com/kaz-yos/distributed
## devtools::install_github(repo = "kaz-yos/distributed")
library(distributed)
```

## Generate data
The following function can be used to construct a single dataset with treated and untreated with confounding.
```
     GenerateOneCenter(n, AssignCovariates, alphas, betas, survParams)

Arguments:

       n: study site-specific sample size

AssignCovariates: covariate generation functions that takes n and p as
          the only arguments.

  alphas: parameter vector for treatment model including c(alpha0,
          alphaX)

   betas: parameter vector for outcome model shared among binary and
          survival outcome models including ‘c(beta0, betaX, betaA,
          betaXA)’.

survParams: vector of two. The first element is the baseline hazard of
          events in the exponential event time outcome model
          (‘lambda’). The second element is the baseline hazard of
          censoring in the exponential censoring time model
          (‘lambda_c’).
```
This function call produces both binary and survival outcome. We focus on the binary outcome.
```{r}
data1 <-
    GenerateOneCenter(n = 1000,
                      AssignCovariates = AssignCovariatesNormBinDefault,
                      alphas = c(alpha0 = -0.5, alphaX = c(0.5, 0.5)),
                      betas = c(beta0 = -0.5, betaX = c(0.5, 0.5),
                                ## Protective effect
                                betaA = -0.3,
                                ## No effect (measure) modification
                                betaXA = c(0, 0)),
                      survParams = c(lambda = -log(0.95), lambda_c = -log(0.99), Tmax = 1))
```
The true data including the counterfactual quantities can be displayed. The overall difference in pY0 (mean probability of response under no treatment) and pY1 (mean probability of response under treatment) is the average treatment effect (ATE). The same thing in the treated (A=1) is the average treatment effect on the treated (ATT). The treatment group difference in pY0 indicates difference in the mean counterfactuals under no treatment. That is, confounding. Similar confounding exhibits in pY1.
```{r}
summary(data1, truth = TRUE)
```

--------------------
- Top Page: http://rpubs.com/kaz_yos/
- Github: https://github.com/kaz-yos

```{r}
print(sessionInfo())
## Record execution time and multicore use
end_time <- Sys.time()
diff_time <- difftime(end_time, start_time, units = "auto")
cat("Started  ", as.character(start_time), "\n",
    "Finished ", as.character(end_time), "\n",
    "Time difference of ", diff_time, " ", attr(diff_time, "units"), "\n",
    "Used ", foreach::getDoParWorkers(), " cores\n",
    "Used ", foreach::getDoParName(), " as backend\n",
    sep = "")
```
