---
title: "BIDA in Stan: Piecewise constant hazard Cox"
author: "Kazuki Yoshida"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: html_document
---

```{r, message = FALSE, tidy = FALSE, echo = F}
## knitr configuration: http://yihui.name/knitr/options#chunk_options
library(knitr)
showMessage <- FALSE
showWarning <- TRUE
set_alias(w = "fig.width", h = "fig.height", res = "results")
opts_chunk$set(comment = "##", error= TRUE, warning = showWarning, message = showMessage,
               tidy = FALSE, cache = F, echo = T,
               fig.width = 7, fig.height = 7, dev.args = list(family = "sans"))
## for rgl
## knit_hooks$set(rgl = hook_rgl, webgl = hook_webgl)
## for animation
opts_knit$set(animation.fun = hook_ffmpeg_html)

## R configuration
options(width = 116, scipen = 5)
```

## References
- Online
  - [Stan Leukemia example](https://github.com/stan-dev/example-models/blob/master/bugs_examples/vol1/leuk/leuk.stan)
  - [Stan for survival models](https://discourse.mc-stan.org/t/stan-for-survival-models/4146)
  - [PyMC3 Bayesian Survival Analysis](https://docs.pymc.io/notebooks/survival_analysis.html)
  - [Piece-Wise Exponential Model](http://data.princeton.edu/wws509/notes/c7s4.html)
- Books
  - [(BIDA) Bayesian Ideas and Data Analysis An Introduction for Scientists and Statisticians](http://blogs.oregonstate.edu/bida/)
  - [(BUGS) The BUGS Book: A Practical Introduction to Bayesian Analysis](https://www.mrc-bsu.cam.ac.uk/software/bugs/the-bugs-project-the-bugs-book/)
  - [(BSA) Bayesian Survival Analysis](https://www.springer.com/us/book/9780387952772)


## Background

The proportional hazards (PH) model (Cox 1972) is a very popular regression method for survival data. The popularity likely originated in not having to specify the (baseline) hazard function. These days, it is popular probably because it is the only survival regression model that many applied researchers are aware of. In Frequentist PH model, parameter estimation is conducted via partial likelihood from which the baseline hazard function has dropped out (BSA p16).

In Bayesian paradigm, obtaining the posterior distribution of parameters requires the full likelihood function involving all parameters including nuisance ones and the prior for all these parameters. In the case of survival analysis, one of the parameters that requires modeling is the entire baseline hazard function. One way to proceed is to parametrize the baseline hazard function parsimoniously, i.e., parametric survival analysis (BIDA p325, BSA chap 2). The other approach is to more flexibly model the baseline hazard function.

Here we will examine the simplest form of the latter, the piecewise constant hazard model.

## Piecewise constant hazard model

This model formulation was taken from BSA (p47-).

Firstly, partition the time axis into $J$ intervals using $0 < s_{1} < s_{2} ... < s_{J} < \infty$.

 $$(0,s_{1}], (s_{1},s_{2}], ..., (s_{J-1},s_{J}]$$

$s_{J}$ is a finite value that has to be larger than the largest observed time in the study. Name each interval $I_{j} = (s_{j-1},s_{j}]$. We assume a constant hazard $\lambda$ within each interval. Let $D = (n,\mathbf{y}, X, \nu)$ describe the observed data.

$\mathbf{y} = (y_{1}, ..., y_{n})^{T}$ is the observed times.

$X$ is a $n \times p$ matrix of covariates associated with a length $p$ vector $\boldsymbol{\beta}$.

$\boldsymbol{\nu} = (\nu_{1},...,\nu_{n})^{T}$ is a vector of failure (censoring) indicators.

Let $\boldsymbol{\lambda} = (\lambda_{1},...,\lambda_{J})^{T}$. The full likelihood for $(\boldsymbol{\beta}, \boldsymbol{\lambda})$ is the following.

$$L(\boldsymbol{\beta},\boldsymbol{\lambda} | D) = \prod^{n}_{i=1} \prod^{J}_{j=1} (\lambda_{j} \exp(\mathbf{x}_{i}^{T}\boldsymbol{\beta}))^{\delta_{ij}\nu_{i}} \exp \left\{ - \delta_{ij} \left[ \lambda_{j}(y_{i} - s_{j-1}) + \sum^{j-1}_{g=1} \lambda_{g}(s_{g} - s_{g-1}) \right] \exp(\mathbf{x}_{i}^{T}\boldsymbol{\beta}) \right\}$$

where $\delta_{ij}$ is an interval specific indicator of end of follow up indicator $I[y_{i} \in I_{j}]$.

This is really hard, so let us dissect this into more manageable pieces. Firstly, we will focus on one individual rather than the entire dataset.

$$L(\boldsymbol{\beta},\boldsymbol{\lambda} | D_{i}) = \prod^{J}_{j=1} (\lambda_{j} \exp(\mathbf{x}_{i}^{T}\boldsymbol{\beta}))^{\delta_{ij}\nu_{i}} \exp \left\{ - \delta_{ij} \left[ \lambda_{j}(y_{i} - s_{j-1}) + \sum^{j-1}_{g=1} \lambda_{g}(s_{g} - s_{g-1}) \right] \exp(\mathbf{x}_{i}^{T}\boldsymbol{\beta}) \right\}$$

Thus, this is a product of interval-specific contributions. However, for interval $I_{j}$ in which the individual just survived without death or censoring $\delta_{ij} = 0$, there is no contribution (no parameters left).

$$(\lambda_{j} \exp(\mathbf{x}_{i}^{T}\boldsymbol{\beta}))^{(0) \nu_{i}} \exp \left\{ - (0) \left[ \lambda_{j}(y_{i} - s_{j-1}) + \sum^{j-1}_{g=1} \lambda_{g}(s_{g} - s_{g-1}) \right] \exp(\mathbf{x}_{i}^{T}\boldsymbol{\beta}) \right\} = 1$$

Therefore, the only contribution happens at the interval when the individual either dies or becomes censored. Note in general, an event individual contributes a density, which is a product of hazard and survival. On the other hand, a censored individual can only contribute survival information.


### Event individual contribution

Firstly, we will consider death in interval $I_{j}$.

$$(\lambda_{j} \exp(\mathbf{x}_{i}^{T}\boldsymbol{\beta}))^{(1) (1)} \exp \left\{ - (1) \left[ \lambda_{j}(y_{i} - s_{j-1}) + \sum^{j-1}_{g=1} \lambda_{g}(s_{g} - s_{g-1}) \right] \exp(\mathbf{x}_{i}^{T}\boldsymbol{\beta}) \right\}$$

$(\lambda_{j} \exp(\mathbf{x}_{i}^{T}\boldsymbol{\beta}))^{(1) (1)}$ is the hazard contribution.

$\exp \left\{ - (1) \left[ \lambda_{j}(y_{i} - s_{j-1}) + \sum^{j-1}_{g=1} \lambda_{g}(s_{g} - s_{g-1}) \right] \exp(\mathbf{x}_{i}^{T}\boldsymbol{\beta}) \right\}$ is the survival contribution. Taken together this individual contributes density information.

Note $\left[ \lambda_{j}(y_{i} - s_{j-1}) + \sum^{j-1}_{g=1} \lambda_{g}(s_{g} - s_{g-1}) \right]$ is the cumulative baseline hazard that this individual faced. By multiplying this with time-constant (by the PH assumption) multiplication factor, $\exp(\mathbf{x}_{i}^{T}\boldsymbol{\beta})$, we obtain the individual-specific cumulative hazard. An exponential of the negative cumulative hazard is the survival.

### Censored individual contribution

Now we will consider censoring in interval $I_{j}$.

$$(\lambda_{j} \exp(\mathbf{x}_{i}^{T}\boldsymbol{\beta}))^{(1) (0)} \exp \left\{ - (1) \left[ \lambda_{j}(y_{i} - s_{j-1}) + \sum^{j-1}_{g=1} \lambda_{g}(s_{g} - s_{g-1}) \right] \exp(\mathbf{x}_{i}^{T}\boldsymbol{\beta}) \right\}$$

$(\lambda_{j} \exp(\mathbf{x}_{i}^{T}\boldsymbol{\beta}))^{(1) (0)} = 1$. There is no hazard contribution.

$\exp \left\{ - (1) \left[ \lambda_{j}(y_{i} - s_{j-1}) + \sum^{j-1}_{g=1} \lambda_{g}(s_{g} - s_{g-1}) \right] \exp(\mathbf{x}_{i}^{T}\boldsymbol{\beta}) \right\}$ is the survival contribution.

### Poisson trick

Let us examine the likelihood for individual $i$ further. As stated above only contribution happens at the interval in which follow up ends by death or censoring. Without loss of generality, consider this interval as $I_{j}$.

$$
\begin{align*}
L(\boldsymbol{\beta},\boldsymbol{\lambda} | D_{i})
&= (\lambda_{j} \exp(\mathbf{x}_{i}^{T}\boldsymbol{\beta}))^{\delta_{ij}\nu_{i}} \exp \left\{ - \delta_{ij} \left[ \lambda_{j}(y_{i} - s_{j-1}) + \sum^{j-1}_{g=1} \lambda_{g}(s_{g} - s_{g-1}) \right] \exp(\mathbf{x}_{i}^{T}\boldsymbol{\beta}) \right\}\\
 &~~~\text{By } \delta_{ij} = 1\\
 &= (\lambda_{j} \exp(\mathbf{x}_{i}^{T}\boldsymbol{\beta}))^{\nu_{i}} \exp \left\{ - \left[ \lambda_{j}(y_{i} - s_{j-1}) + \sum^{j-1}_{g=1} \lambda_{g}(s_{g} - s_{g-1}) \right] \exp(\mathbf{x}_{i}^{T}\boldsymbol{\beta}) \right\}\\
 &~~~\text{Let $H_{i,g}$ represent at-risk time during $I_{g}$ for $i$}.\\
 &= (\lambda_{j} \exp(\mathbf{x}_{i}^{T}\boldsymbol{\beta}))^{\nu_{i}} \exp \left\{ - \left[ \sum^{j}_{g=1} \lambda_{g} H_{i,g} \right] \exp(\mathbf{x}_{i}^{T}\boldsymbol{\beta}) \right\}\\
 &= (\lambda_{j} \exp(\mathbf{x}_{i}^{T}\boldsymbol{\beta}))^{\nu_{i}} \exp \left\{ - \sum^{j}_{g=1} \lambda_{g} H_{i,g} \exp(\mathbf{x}_{i}^{T}\boldsymbol{\beta}) \right\}\\
 &= (\lambda_{j} \exp(\mathbf{x}_{i}^{T}\boldsymbol{\beta}))^{\nu_{i}} \prod^{j}_{g=1} \exp \left\{ - \lambda_{g} H_{i,g} \exp(\mathbf{x}_{i}^{T}\boldsymbol{\beta}) \right\}\\
 &~~~\text{Note the new middle piece is 1.}\\
 &= (\lambda_{j} \exp(\mathbf{x}_{i}^{T}\boldsymbol{\beta}))^{\nu_{i}} \left[ \prod^{j-1}_{g=1} (\lambda_{g} H_{i,g} \exp(\mathbf{x}_{i}^{T}\boldsymbol{\beta}))^{0} \right] \prod^{j}_{g=1} \exp \left\{ - \lambda_{g} H_{i,g} \exp(\mathbf{x}_{i}^{T}\boldsymbol{\beta}) \right\}\\
 &\propto (\lambda_{j} H_{i,j} \exp(\mathbf{x}_{i}^{T}\boldsymbol{\beta}))^{\nu_{i}} \left[ \prod^{j-1}_{g=1} (\lambda_{g} H_{i,g} \exp(\mathbf{x}_{i}^{T}\boldsymbol{\beta}))^{0} \right] \prod^{j}_{g=1} \exp \left\{ - \lambda_{g} H_{i,g} \exp(\mathbf{x}_{i}^{T}\boldsymbol{\beta}) \right\}\\
 &~~~\text{By reintroducing } \delta_{i,g}\\
 &= \left[ \prod^{j}_{g=1} (\lambda_{g} H_{i,g} \exp(\mathbf{x}_{i}^{T}\boldsymbol{\beta}))^{\delta_{i,g} \nu_{i}} \right] \prod^{j}_{g=1} \exp \left\{ - \lambda_{g} H_{i,g} \exp(\mathbf{x}_{i}^{T}\boldsymbol{\beta}) \right\}\\
 &= \prod^{j}_{g=1} \exp \left\{ - \lambda_{g} H_{i,g} \exp(\mathbf{x}_{i}^{T}\boldsymbol{\beta}) \right\} (\lambda_{g} H_{i,g} \exp(\mathbf{x}_{i}^{T}\boldsymbol{\beta}))^{\delta_{i,g} \nu_{i}}\\
 &~~~\text{By } \delta_{i,g} \nu_{i} \in \{0,1\}\\
 &= \prod^{j}_{g=1} \exp \left\{ - \lambda_{g} H_{i,g} \exp(\mathbf{x}_{i}^{T}\boldsymbol{\beta}) \right\} (\lambda_{g} H_{i,g} \exp(\mathbf{x}_{i}^{T}\boldsymbol{\beta}))^{\delta_{i,g} \nu_{i}} \big/ (\delta_{i,g} \nu_{i})!\\
\end{align*}
$$

We can multiply the likelihood with a term that does not contain parameters and retain the same inference. Note the last expression is a product of individual- and interval-specific Poisson likelihood. This transformation implies that we can split each individual's observation into interval-specific observations up until the interval in which follow up ended. In addition to copying covariates, each interval-specific observation has to have the duration of the at-risk time and an indicator that is 1 in the last interval if the individual died otherwise 0. The latter indicator serves as the outcome of the Poisson model.

The corresponding Poisson model for individual $i$ interval $g$ is the following.

$$
\begin{align*}
\mu_{i,g} &= \lambda_{g} H_{i,g} \exp(\mathbf{x}_{i}^{T}\boldsymbol{\beta})\\
\log(\mu_{i,g}) &= \log(\lambda_{g} H_{i,g} \exp(\mathbf{x}_{i}^{T}\boldsymbol{\beta}))\\
 &= \log(H_{i,g}) + \log(\lambda_{g}) + \mathbf{x}_{i}^{T}\boldsymbol{\beta}\\
 &= \log(H_{i,g}) + \beta_{0,g} + \mathbf{x}_{i}^{T}\boldsymbol{\beta}\\
\end{align*}
$$

Therefore, $\log(H_{i,g})$ becomes the offset. The intercept $\beta_{0,g} = \log(\lambda_{g})$ is interval-specific.

## Load packages

```{r}
library(tidyverse)
```

##

```{r}

```


--------------------
- Top Page: http://rpubs.com/kaz_yos/
- Github: https://github.com/kaz-yos
