---
title: "BIDA: Density Estimation (Finite Mixture Model)"
author: "Kazuki Yoshida"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: html_document
---

```{r, message = FALSE, tidy = FALSE, echo = F}
## knitr configuration: http://yihui.name/knitr/options#chunk_options
library(knitr)
showMessage <- FALSE
showWarning <- TRUE
set_alias(w = "fig.width", h = "fig.height", res = "results")
opts_chunk$set(comment = "##", error= TRUE, warning = showWarning, message = showMessage,
               tidy = FALSE, cache = FALSE, echo = TRUE,
               fig.width = 7, fig.height = 7, dev.args = list(family = "sans"))
## for rgl
## knit_hooks$set(rgl = hook_rgl, webgl = hook_webgl)
## for animation
opts_knit$set(animation.fun = hook_ffmpeg_html)
## R configuration
options(width = 116, scipen = 5)
## Record start time
start_time <- Sys.time()
## Configure parallelization
## Parallel backend for foreach (also loads foreach and parallel; includes doMC)
library(doParallel)
## Reproducible parallelization
library(doRNG)
## Detect core count (Do not use on clusters)
n_cores <- parallel::detectCores()
## Used by parallel::mclapply() as default
options(mc.cores = n_cores)
## Used by doParallel as default
options(cores = n_cores)
## Register doParallel as the parallel backend for foreach
## http://stackoverflow.com/questions/28989855/the-difference-between-domc-and-doparallel-in-r
doParallel::registerDoParallel(cores = n_cores)
set.seed(165635606)
```

## References
- Books
  - [(BIDA) Bayesian Ideas and Data Analysis An Introduction for Scientists and Statisticians](http://blogs.oregonstate.edu/bida/)
  - [(BDA) Bayesian Data Analysis](http://www.stat.columbia.edu/~gelman/book/)
- Web sites
  - [Bayesian Statistics Using Stan 10 Finite Mixtures](http://www.stat.columbia.edu/~gelman/bda.course/_book/mixture-modeling-chapter.html)
  - [Finite mixture models in Stan](http://modernstatisticalworkflow.blogspot.com/2016/10/finite-mixture-models-in-stan.html)
  - [Identifying Bayesian Mixture Models](http://mc-stan.org/users/documentation/case-studies/identifying_mixture_models.html)
  - [Taming Divergences in Stan Models](https://www.martinmodrak.cz/2018/02/19/taming-divergences-in-stan-models/)
  - [Check HMC diagnostics after sampling](http://mc-stan.org/rstan/reference/check_hmc_diagnostics.html)
  - [Mixture models in Stan: you can use log_mix](https://andrewgelman.com/2017/08/21/mixture-models-stan-can-use-log_mix/)
  - [Identification of mixture of multivariate normal distributions](https://discourse.mc-stan.org/t/identification-of-mixture-of-multivariate-normal-distributions/4203)
- jSoftware
  - [Stan Modeling Language User's Guide and Reference Manual, Version 2.17.0](http://mc-stan.org/users/documentation/)
  - [CRAN DPpackage: Bayesian Nonparametric Modeling in R](https://cran.r-project.org/package=DPpackage)
  - [J Stat Softw. 2011. DPpackage: Bayesian Non- and Semi-parametric Modelling in R.](https://www.jstatsoft.org/article/view/v040i05)


## Background
Here we have 82 data points in the galaxy from a 1-dimensional unknown distribution. The aim is to fit a normal finite mixture model with a pre-specified number of latent clusters.

## Load packages

```{r}
library(tidyverse)
library(rstan)
## devtools::install_github('jburos/biostan', build_vignettes = TRUE, dependencies = TRUE)
## library(biostan)
library(DPpackage)
set.seed(732565397)
```

## Prepare data
```{r}
data(galaxy, package = "DPpackage")
galaxy <- galaxy %>%
    as_data_frame() %>%
    mutate(log_speed = log(speed),
           k_speed = speed / 1000)
galaxy
ggplot(data = galaxy, mapping = aes(x = k_speed)) +
    geom_point(y = 0.5) +
    scale_y_continuous(limits = c(0,1), breaks = NULL) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
          legend.key = element_blank(),
          plot.title = element_text(hjust = 0.5),
          strip.background = element_blank())
```

The speed data seem to show some distinct clusters. We will use the following grid for later visualization.

```{r}
grid_max <- 40
grid_min <- -20
grid_length <- 100
```

## Define helper functions
Helper functions for here and later use.
```{r}
print_relevant_pars <- function(fit, pars = c("mu","sigma","Pi","lp__")) {
    print(fit, pars = pars)
}

traceplot_all <- function(fit, pars = c("mu","sigma","Pi","lp__")) {
    for (par in pars) {
        print(traceplot(fit, inc_warmup = TRUE, pars = par))
    }
}

pairs_plot_all <- function(fit, pars = c("mu","sigma","Pi")) {
    for (par in pars) {
        pairs(fit, pars = par)
    }
}

plot_draws <- function(stan_fit) {
    ## Note direct access to global variables
    draw_data  <- tidybayes::tidy_draws(stan_fit) %>%
        select(.chain, .iteration, .draw, starts_with("log_f")) %>%
        gather(key = key, value = value, starts_with("log_f")) %>%
        mutate(key = gsub("log_f|\\[|\\]", "", key) %>% as.integer(),
           x = factor(key, labels = seq(from = grid_min, to = grid_max, length.out = grid_length)) %>%
               as.character() %>%
               as.numeric(),
           value = exp(value))

    summary_density <- draw_data %>%
        group_by(.chain, x) %>%
        summarize(value = mean(value))

    ggplot(data = draw_data, mapping = aes(x = x, y = value,
           group = interaction(.chain, .iteration, .draw))) +
    ## geom_line(size = 0.1, alpha = 1/20) +
    geom_line(data = summary_density, mapping = aes(group = .chain), size = 0.5, color = "gray") +
    geom_point(data = galaxy, mapping = aes(x = k_speed, group = NULL), y = 0) +
    facet_wrap(~ .chain) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
          legend.key = element_blank(),
          plot.title = element_text(hjust = 0.5),
          strip.background = element_blank())
}
```

## Finite Mixture of Normals (Unordered mu)
#### 3 latent cluster model
```{r, results = 'hide'}
H <- 3
normal_fixed_mixture_stan_tau3 <-
    rstan::stan(model_code = readr::read_file("./bayesianideas_density_normal_fixed_mixture_tau_unordered.stan"),
                data = list(
                    ## tau ~ Gamma(alpha, beta)
                    alpha = rep(10^(1), H), beta = rep(10^(0), H),
                    ## mu ~ N(m, s_squared)
                    m = c(20,20,20), s_squared = c(4,4,4),
                    ## Pi ~ Dirichlet(dirichlet_alpha / H)
                    dirichlet_alpha = c(1,1,1) * 3 * 3,
                    ## Number of clusters
                    H = H,
                    n = nrow(galaxy),
                    y = galaxy$k_speed,
                    grid_max = grid_max,
                    grid_min = grid_min,
                    grid_length = grid_length),
                iter = 2000*5,
                chains = 12*3)
```
```{r}
print_relevant_pars(normal_fixed_mixture_stan_tau3)
traceplot_all(normal_fixed_mixture_stan_tau3)
pairs_plot_all(normal_fixed_mixture_stan_tau3)
pairs(normal_fixed_mixture_stan_tau3, pars = c("mu","sigma","Pi"))
plot_draws(normal_fixed_mixture_stan_tau3)
## shinystan::launch_shinystan(normal_fixed_mixture_stan_tau3)
```

#### 6 latent cluster model
```{r, results = 'hide'}
H <- 6
normal_fixed_mixture_stan_tau6 <-
    rstan::stan(model_code = readr::read_file("./bayesianideas_density_normal_fixed_mixture_tau_unordered.stan"),
                data = list(
                    ## tau ~ Gamma(alpha, beta)
                    alpha = rep(10^(1), H), beta = rep(10^(0), H),
                    ## mu ~ N(m, s_squared)
                    m = c(20,20,20), s_squared = c(4,4,4),
                    ## Pi ~ Dirichlet(dirichlet_alpha / H)
                    dirichlet_alpha = c(1,1,1) * 3 * 3,
                    ## Number of clusters
                    H = H,
                    n = nrow(galaxy),
                    y = galaxy$k_speed,
                    grid_max = grid_max,
                    grid_min = grid_min,
                    grid_length = grid_length),
                iter = 2000*5,
                chains = 12*3)
```
```{r}
print_relevant_pars(normal_fixed_mixture_stan_tau6)
traceplot_all(normal_fixed_mixture_stan_tau6)
pairs_plot_all(normal_fixed_mixture_stan_tau6)
pairs(normal_fixed_mixture_stan_tau6, pars = c("mu","sigma","Pi"))
plot_draws(normal_fixed_mixture_stan_tau6)
## shinystan::launch_shinystan(normal_fixed_mixture_stan_tau3)
```

## Finite Mixture of Normals (Ordered mu)
#### 3 latent cluster model
```{r, results = 'hide'}
H <- 3
normal_fixed_mixture_stan_tau3 <-
    rstan::stan(model_code = readr::read_file("./bayesianideas_density_normal_fixed_mixture_tau.stan"),
                data = list(
                    ## tau ~ Gamma(alpha, beta)
                    alpha = rep(10^(1), H), beta = rep(10^(0), H),
                    ## mu ~ N(m, s_squared)
                    m = c(10,20,30), s_squared = c(4,4,4),
                    ## Pi ~ Dirichlet(dirichlet_alpha / H)
                    dirichlet_alpha = c(1,3,1) * 3,
                    ## Number of clusters
                    H = H,
                    n = nrow(galaxy),
                    y = galaxy$k_speed,
                    grid_max = grid_max,
                    grid_min = grid_min,
                    grid_length = grid_length),
                iter = 2000*5,
                chains = 12*3)
```
```{r}
print_relevant_pars(normal_fixed_mixture_stan_tau3)
traceplot_all(normal_fixed_mixture_stan_tau3)
pairs_plot_all(normal_fixed_mixture_stan_tau3)
pairs(normal_fixed_mixture_stan_tau3, pars = c("mu","sigma","Pi"))
plot_draws(normal_fixed_mixture_stan_tau3)
## shinystan::launch_shinystan(normal_fixed_mixture_stan_tau3)
```

#### 6 latent cluster model
```{r, results = 'hide'}
H <- 6
normal_fixed_mixture_stan_tau6 <-
    rstan::stan(model_code = readr::read_file("./bayesianideas_density_normal_fixed_mixture_tau.stan"),
                data = list(
                    ## tau ~ Gamma(alpha, beta)
                    alpha = rep(10^(1), H), beta = rep(10^(0), H),
                    ## mu ~ N(m, s_squared)
                    m = c(10,20,30), s_squared = c(4,4,4),
                    ## Pi ~ Dirichlet(dirichlet_alpha / H)
                    dirichlet_alpha = c(1,3,1) * 3,
                    ## Number of clusters
                    H = H,
                    n = nrow(galaxy),
                    y = galaxy$k_speed,
                    grid_max = grid_max,
                    grid_min = grid_min,
                    grid_length = grid_length),
                iter = 2000*5,
                chains = 12*3)
```
```{r}
print_relevant_pars(normal_fixed_mixture_stan_tau6)
traceplot_all(normal_fixed_mixture_stan_tau6)
pairs_plot_all(normal_fixed_mixture_stan_tau6)
pairs(normal_fixed_mixture_stan_tau6, pars = c("mu","sigma","Pi"))
plot_draws(normal_fixed_mixture_stan_tau6)
## shinystan::launch_shinystan(normal_fixed_mixture_stan_tau6)
```


--------------------
- Top Page: http://rpubs.com/kaz_yos/
- Github: https://github.com/kaz-yos

```{r}
print(sessionInfo())
## Record execution time and multicore use
end_time <- Sys.time()
diff_time <- difftime(end_time, start_time, units = "auto")
cat("Started  ", as.character(start_time), "\n",
    "Finished ", as.character(end_time), "\n",
    "Time difference of ", diff_time, " ", attr(diff_time, "units"), "\n",
    "Used ", foreach::getDoParWorkers(), " cores\n",
    "Used ", foreach::getDoParName(), " as backend\n",
    sep = "")
```
