---
title: "Bayesian Propensity Score"
author: "Kazuki Yoshida"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: html_document
---

```{r, message = FALSE, tidy = FALSE, echo = F}
## knitr configuration: http://yihui.name/knitr/options#chunk_options
library(knitr)
showMessage <- FALSE
showWarning <- TRUE
set_alias(w = "fig.width", h = "fig.height", res = "results")
opts_chunk$set(comment = "##", error= TRUE, warning = showWarning, message = showMessage,
               tidy = FALSE, cache = FALSE, echo = TRUE,
               fig.width = 7, fig.height = 7, dev.args = list(family = "sans"))
## for rgl
## knit_hooks$set(rgl = hook_rgl, webgl = hook_webgl)
## for animation
opts_knit$set(animation.fun = hook_ffmpeg_html)
## R configuration
options(width = 116, scipen = 5)
## Record start time
start_time <- Sys.time()
## Configure parallelization
## Parallel backend for foreach (also loads foreach and parallel; includes doMC)
library(doParallel)
## Reproducible parallelization
library(doRNG)
## Detect core count (Do not use on clusters)
n_cores <- parallel::detectCores()
## Used by parallel::mclapply() as default
options(mc.cores = n_cores)
## Used by doParallel as default
options(cores = n_cores)
## Register doParallel as the parallel backend for foreach
## http://stackoverflow.com/questions/28989855/the-difference-between-domc-and-doparallel-in-r
doParallel::registerDoParallel(cores = n_cores)
```

## References
- Papers
  - [Hoshino 2008. A Bayesian Propensity Score Adjustment for Latent Variable Modeling and MCMC Algorithm](https://www.researchgate.net/publication/223577619_A_Bayesian_propensity_score_adjustment_for_latent_variable_modeling_and_MCMC_algorithm)
  - [McCandless 2009. ]()
  - [McCandless 2010. ]()
  - [Zigler 2017. Posterior Predictive Treatment Assignment for Estimating Causal Effects with Limited Overlap](https://arxiv.org/abs/1710.08749)
- Stan
  - [Stan (+R) Workshop 2019 at Calvin College](https://rpruim.github.io/StanWorkshop/)
  - [An easy way to simulate fake data from your Stan model](http://modernstatisticalworkflow.blogspot.com/2017/04/an-easy-way-to-simulate-fake-data-from.html?m=1)


## Load packages
```{r}
## random.org
set.seed(842815442)
library(tidyverse)
## Simulation suite from
## https://github.com/kaz-yos/distributed
## devtools::install_github(repo = "kaz-yos/distributed")
library(distributed)
library(rstan)
library(bayesplot)
```

## Existing methods
Multiple papers have proposed various Bayesian propensity score methods.

| Paper           | Type   | Model 1 | PS Use    | Model 2    | Estimand    |
|-----------------|--------|---------|-----------|------------|-------------|
| Hoshino 2008    | Joint  | A~L     | Weight    |            |             |
| McCandless 2009 | Joint  | A~L     | Strata    | Y~A,Strata | Conditional |
| McCandless 2010 | 2-step | A~L     | Adjust    | Y~A,PS     | Conditional |
| An 2010         | Joint  |         | Adj,Match |            |             |
| Kaplan 2012     | 2-step |         |           |            |             |
| Zigler 2013     |        |         |           |            |             |
| Chen 2014       | 2-step |         |           |            |             |
| Saarela 2015    |        |         |           |            |             |
| Saarela 2016    |        |         |           |            |             |
| Zigler 2016     |        |         |           |            |             |
| Zigler 2017     |        |         |           |            | Cond? ATO   |
| Capistrano 2019 |        |         |           |            | ATE         |
| Liao 2019       |        |         |           |            |             |

--------------------
- Top Page: http://rpubs.com/kaz_yos/
- Github: https://github.com/kaz-yos

```{r}
print(sessionInfo())
## Record execution time and multicore use
end_time <- Sys.time()
diff_time <- difftime(end_time, start_time, units = "auto")
cat("Started  ", as.character(start_time), "\n",
    "Finished ", as.character(end_time), "\n",
    "Time difference of ", diff_time, " ", attr(diff_time, "units"), "\n",
    "Used ", foreach::getDoParWorkers(), " cores\n",
    "Used ", foreach::getDoParName(), " as backend\n",
    sep = "")
```
