---
title: "Beta Regression in Stan"
author: "Kazuki Yoshida"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: html_document
---

```{r, message = FALSE, tidy = FALSE, echo = F}
## knitr configuration: http://yihui.name/knitr/options#chunk_options
library(knitr)
showMessage <- FALSE
showWarning <- TRUE
set_alias(w = "fig.width", h = "fig.height", res = "results")
opts_chunk$set(comment = "##", error= TRUE, warning = showWarning, message = showMessage,
               tidy = FALSE, cache = FALSE, echo = TRUE,
               fig.width = 7, fig.height = 7, dev.args = list(family = "sans"))
## for rgl
## knit_hooks$set(rgl = hook_rgl, webgl = hook_webgl)
## for animation
opts_knit$set(animation.fun = hook_ffmpeg_html)
## R configuration
options(width = 116, scipen = 5)
## Record start time
start_time <- Sys.time()
## Configure parallelization
## Parallel backend for foreach (also loads foreach and parallel; includes doMC)
library(doParallel)
## Reproducible parallelization
library(doRNG)
## Detect core count (Do not use on clusters)
n_cores <- parallel::detectCores()
## Used by parallel::mclapply() as default
options(mc.cores = n_cores)
## Used by doParallel as default
options(cores = n_cores)
## Register doParallel as the parallel backend for foreach
## http://stackoverflow.com/questions/28989855/the-difference-between-domc-and-doparallel-in-r
doParallel::registerDoParallel(cores = n_cores)
```

## References
- [Extended Beta Regression in R: Shaken, Stirred, Mixed, and Partitioned](https://www.jstatsoft.org/article/view/v048i11)
- [Beta distribution (Wikipedia)](https://en.wikipedia.org/wiki/Beta_distribution)

## Load packages

```{r}
library(tidyverse)
library(rstan)
```

## Load data
```
FoodExpenditure            package:betareg             R Documentation
Proportion of Household Income Spent on Food
Description:
     Data on proportion of income spent on food for a random sample of
     38 households in a large US city.
Usage:
     data("FoodExpenditure")
Format:
     A data frame containing 38 observations on 3 variables.
     food household expenditures for food.
     income household income.
     persons number of persons living in household.
Source:
     Taken from Griffiths et al. (1993, Table 15.4).
```
```{r}
data(FoodExpenditure, package = "betareg")
FoodExpenditure <-
    FoodExpenditure %>%
    mutate(p_food = food / income)
FoodExpenditure
```
```{r}
ggplot(data = FoodExpenditure, mapping = aes(x = persons, y = p_food)) +
    geom_point() +
    scale_y_continuous(limits = c(0,1)) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
          legend.key = element_blank(),
          plot.title = element_text(hjust = 0.5),
          strip.background = element_blank())

```

## Model
We will use the mean-precision parametrization of the beta distribution.

$$\begin{align*}
  f(y;\mu,\phi) &= \frac{\Gamma(\phi)}{\Gamma(\mu\phi)\Gamma((1-\mu)\phi)}y^{\mu\phi-1}(1-y)^{(1-\mu)\phi-1}, y \in (0,1)
  \end{align*}
$$

The correspondence to the typical parametrization is: $\alpha = \mu\phi$ and $\beta = (1-\mu)\phi$. As a result the mean and variance are as follows.

$$\begin{align*}
  E[Y] &= \frac{\alpha}{\alpha+\beta}\\
       &= \frac{\mu\phi}{\mu\phi + (1-\mu)\phi}\\
       &= \frac{\mu\phi}{\phi}\\
       &= \mu\\
\\
Var(Y) &= \frac{\alpha\beta}{(\alpha+\beta)^{2}(\alpha+\beta+1)}\\
       &= \frac{\mu\phi(1-\mu)\phi}{\phi^{2}(\phi+1)}\\
       &= \frac{\mu(1-\mu)}{\phi+1}\\
  \end{align*}
$$

These mean and precision parameters can be modeled as functions of covariates.

$$\begin{align*}
  g_{1}(\mu_{i}) &= \eta_{1i} = \mathbb{x}_{1i}^{T}\boldsymbol{\beta}_{x1}\\
  g_{2}(\phi_{i}) &= \eta_{2i} = \mathbb{x}_{2i}^{T}\boldsymbol{\beta}_{x2}\\
  \end{align*}
$$

The sets of covariates $\mathbb{x}_{1i}$ and $\mathbb{x}_{2i}$ can differ.

--------------------
- Top Page: http://rpubs.com/kaz_yos/
- Github: https://github.com/kaz-yos

```{r}
print(sessionInfo())
## Record execution time and multicore use
end_time <- Sys.time()
diff_time <- difftime(end_time, start_time, units = "auto")
cat("Started  ", as.character(start_time), "\n",
    "Finished ", as.character(end_time), "\n",
    "Time difference of ", diff_time, " ", attr(diff_time, "units"), "\n",
    "Used ", foreach::getDoParWorkers(), " cores\n",
    "Used ", foreach::getDoParName(), " as backend\n",
    sep = "")
```
