---
title: "Beta Regression in Stan"
author: "Kazuki Yoshida"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: html_document
---

```{r, message = FALSE, tidy = FALSE, echo = F}
## knitr configuration: http://yihui.name/knitr/options#chunk_options
library(knitr)
showMessage <- FALSE
showWarning <- TRUE
set_alias(w = "fig.width", h = "fig.height", res = "results")
opts_chunk$set(comment = "##", error= TRUE, warning = showWarning, message = showMessage,
               tidy = FALSE, cache = FALSE, echo = TRUE,
               fig.width = 5, fig.height = 5, dev.args = list(family = "sans"))
## for rgl
## knit_hooks$set(rgl = hook_rgl, webgl = hook_webgl)
## for animation
opts_knit$set(animation.fun = hook_ffmpeg_html)
## R configuration
options(width = 116, scipen = 5)
## Record start time
start_time <- Sys.time()
## Configure parallelization
## Parallel backend for foreach (also loads foreach and parallel; includes doMC)
library(doParallel)
## Reproducible parallelization
library(doRNG)
## Detect core count (Do not use on clusters)
n_cores <- parallel::detectCores()
## Used by parallel::mclapply() as default
options(mc.cores = n_cores)
## Used by doParallel as default
options(cores = n_cores)
## Register doParallel as the parallel backend for foreach
## http://stackoverflow.com/questions/28989855/the-difference-between-domc-and-doparallel-in-r
doParallel::registerDoParallel(cores = n_cores)
```

## References
- [Extended Beta Regression in R: Shaken, Stirred, Mixed, and Partitioned](https://www.jstatsoft.org/article/view/v048i11)
- [Beta distribution (Wikipedia)](https://en.wikipedia.org/wiki/Beta_distribution)

## Load packages

```{r}
library(tidyverse)
## https://github.com/jburos/biostan
## devtools::install_github('jburos/biostan', build_vignettes=TRUE, dependencies=TRUE)
library(biostan)
library(rstan)
```

## Load data
```
FoodExpenditure            package:betareg             R Documentation
Proportion of Household Income Spent on Food
Description:
     Data on proportion of income spent on food for a random sample of
     38 households in a large US city.
Usage:
     data("FoodExpenditure")
Format:
     A data frame containing 38 observations on 3 variables.
     food household expenditures for food.
     income household income.
     persons number of persons living in household.
Source:
     Taken from Griffiths et al. (1993, Table 15.4).
```
```{r}
data(FoodExpenditure, package = "betareg")
FoodExpenditure <-
    FoodExpenditure %>%
    as_tibble() %>%
    mutate(p_food = food / income)
FoodExpenditure
```
```{r}
ggplot(data = FoodExpenditure, mapping = aes(x = persons, y = p_food)) +
    geom_point() +
    scale_y_continuous(limits = c(0,1)) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
          legend.key = element_blank(),
          plot.title = element_text(hjust = 0.5),
          strip.background = element_blank())
ggplot(data = FoodExpenditure, mapping = aes(x = income, y = p_food)) +
    geom_point() +
    scale_y_continuous(limits = c(0,1)) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
          legend.key = element_blank(),
          plot.title = element_text(hjust = 0.5),
          strip.background = element_blank())
```


## Model specification
We will use the mean-precision parametrization of the beta distribution.

$$\begin{align*}
  f(y;\mu,\phi) &= \frac{\Gamma(\phi)}{\Gamma(\mu\phi)\Gamma((1-\mu)\phi)}y^{\mu\phi-1}(1-y)^{(1-\mu)\phi-1}, y \in (0,1)
  \end{align*}
$$

The correspondence to the typical parametrization is: $\alpha = \mu\phi$ and $\beta = (1-\mu)\phi$. As a result the mean and variance are as follows.

$$\begin{align*}
  E[Y] &= \frac{\alpha}{\alpha+\beta}\\
       &= \frac{\mu\phi}{\mu\phi + (1-\mu)\phi}\\
       &= \frac{\mu\phi}{\phi}\\
       &= \mu\\
\\
Var(Y) &= \frac{\alpha\beta}{(\alpha+\beta)^{2}(\alpha+\beta+1)}\\
       &= \frac{\mu\phi(1-\mu)\phi}{\phi^{2}(\phi+1)}\\
       &= \frac{\mu(1-\mu)}{\phi+1}\\
  \end{align*}
$$

These mean and precision parameters can be modeled as functions of covariates.

$$\begin{align*}
  g_{1}(\mu_{i}) &= \eta_{1i} = \mathbb{x}_{1i}^{T}\boldsymbol{\beta}_{x1}\\
  g_{2}(\phi_{i}) &= \eta_{2i} = \mathbb{x}_{2i}^{T}\boldsymbol{\beta}_{x2}\\
  \end{align*}
$$

The sets of covariates $\mathbb{x}_{1i}$ and $\mathbb{x}_{2i}$ can differ.

```{r}
stan_logit_log_beta_model_file <- "./beta_regression.stan"
biostan::print_stan_file(stan_logit_log_beta_model_file)
```


## Model fit
```{r}
## Prepare dataset
N <- nrow(FoodExpenditure)
y <- FoodExpenditure$p_food
##  Mean model part
X1 <- model.matrix(object = ~ income + persons, data = FoodExpenditure)
X1_dim <- ncol(X1)
beta_x1_mean <- rep(0, X1_dim)
beta_x1_sd <- c(10, rep(1, X1_dim-1))
##  Precision model part
X2 <- model.matrix(object = ~ income + persons, data = FoodExpenditure)
X2_dim <- ncol(X2)
beta_x2_mean <- rep(0, X2_dim)
beta_x2_sd <- c(10, rep(1, X2_dim-1))
```
```{r, results = 'hide'}
stan_logit_log_beta_model_fit1 <-
    rstan::stan(file = stan_logit_log_beta_model_file,
                data = list(N = N,
                            y = y,
                            X1 = X1,
                            X1_dim = X1_dim,
                            beta_x1_mean = beta_x1_mean,
                            beta_x1_sd = beta_x1_sd,
                            X2 = X2,
                            X2_dim = X2_dim,
                            beta_x2_mean = beta_x2_mean,
                            beta_x2_sd = beta_x2_sd),
                chains = n_cores,
                cores = n_cores,
                verbose = TRUE)
```


## MCMC diagnostics
```{r}
traceplot(stan_logit_log_beta_model_fit1, pars = c("beta_x1","beta_x2"), inc_warmup = TRUE)
pairs(stan_logit_log_beta_model_fit1, pars = c("beta_x1","beta_x2","lp__"))
## More comprehensive and interactive
## shinystan::launch_shinystan(stan_logit_log_beta_model_fit1)
```


## Model results
```{r}
print(stan_logit_log_beta_model_fit1, pars = c("beta_x1","beta_x2","lp__"))
plot(stan_logit_log_beta_model_fit1, pars = c("beta_x1","beta_x2"))
```


## Posterior predictive distribution
```{r}
stan_logit_log_beta_model_fit1_ppd <-
    stan_logit_log_beta_model_fit1 %>%
    tidybayes::tidy_draws() %>%
    select(.chain, .iteration, .draw, starts_with("y_rep")) %>%
    gather(key = key, value = value, starts_with("y_rep")) %>%
    mutate(key = gsub("y_rep|\\[|\\]", "", key) %>% as.integer())

stan_logit_log_beta_model_fit1_ppd_qs <-
    stan_logit_log_beta_model_fit1_ppd %>%
    group_by(key) %>%
    summarize(`0.025` = quantile(value, 0.025),
              `0.975` = quantile(value, 1-0.025),
              `0.100` = quantile(value, 0.1),
              `0.900` = quantile(value, 1-0.1),
              `0.250` = quantile(value, 0.25),
              `0.750` = quantile(value, 1-0.25))

FoodExpenditure %>%
                   mutate(key = seq_along(food)) %>%
    left_join(stan_logit_log_beta_model_fit1_ppd_qs) %>%
    ggplot(mapping = aes(x = key)) +
    geom_ribbon(mapping = aes(ymin = `0.025`, ymax = `0.975`), alpha = 0.1) +
    geom_ribbon(mapping = aes(ymin = `0.100`, ymax = `0.900`), alpha = 0.2) +
    geom_ribbon(mapping = aes(ymin = `0.250`, ymax = `0.750`), alpha = 0.3) +
    geom_point(mapping = aes(y = p_food)) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
          legend.key = element_blank(),
          plot.title = element_text(hjust = 0.5),
          strip.background = element_blank())

```

--------------------
- Top Page: http://rpubs.com/kaz_yos/
- Github: https://github.com/kaz-yos

```{r}
print(sessionInfo())
## Record execution time and multicore use
end_time <- Sys.time()
diff_time <- difftime(end_time, start_time, units = "auto")
cat("Started  ", as.character(start_time), "\n",
    "Finished ", as.character(end_time), "\n",
    "Time difference of ", diff_time, " ", attr(diff_time, "units"), "\n",
    "Used ", foreach::getDoParWorkers(), " cores\n",
    "Used ", foreach::getDoParName(), " as backend\n",
    sep = "")
```
