---
title: "Poisson Models"
author: "Kazuki Yoshida"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: html_document
---

```{r, message = FALSE, tidy = FALSE, echo = F}
## knitr configuration: http://yihui.name/knitr/options#chunk_options
library(knitr)
showMessage <- FALSE
showWarning <- TRUE
set_alias(w = "fig.width", h = "fig.height", res = "results")
opts_chunk$set(comment = "##", error= TRUE, warning = showWarning, message = showMessage,
               tidy = FALSE, cache = FALSE, echo = TRUE,
               fig.width = 7, fig.height = 7, dev.args = list(family = "sans"))
## for rgl
## knit_hooks$set(rgl = hook_rgl, webgl = hook_webgl)
## for animation
opts_knit$set(animation.fun = hook_ffmpeg_html)
## R configuration
options(width = 116, scipen = 5)
## Record start time
start_time <- Sys.time()
## Configure parallelization
## Parallel backend for foreach (also loads foreach and parallel; includes doMC)
library(doParallel)
## Reproducible parallelization
library(doRNG)
## Detect core count (Do not use on clusters)
n_cores <- parallel::detectCores()
## Used by parallel::mclapply() as default
options(mc.cores = n_cores)
## Used by doParallel as default
options(cores = n_cores)
## Register doParallel as the parallel backend for foreach
## http://stackoverflow.com/questions/28989855/the-difference-between-domc-and-doparallel-in-r
doParallel::registerDoParallel(cores = n_cores)
```

## References
- Web
  - [(DAE) Negative Binomial Regression | R Data Analysis Examples](https://stats.idre.ucla.edu/r/dae/negative-binomial-regression/)
  - [4.A Models for Over-Dispersed Count Data](https://data.princeton.edu/wws509/stata/overdispersion)
  - [Models for Count Data With Overdispersion](https://data.princeton.edu/wws509/notes/c4a.pdf)
  - [An Introduction to ggdag](https://cran.r-project.org/web/packages/ggdag/vignettes/intro-to-ggdag.html)
- Books
  - [Bayesian Models: a Statistical Primer for Ecologists](https://press.princeton.edu/titles/10523.html)

## Load packages
```{r}
library(tidyverse)

```

## Load data
```{r, cache = TRUE}
data1 <- haven::read_dta("https://stats.idre.ucla.edu/stat/stata/dae/nb_data.dta")
```
```{r}
data1 <- data1 %>%
    mutate(prog = factor(prog, levels = 1:3, labels = c("General", "Academic", "Vocational")),
           id = factor(id))
data1
```

## Visualize
```{r}
data1 %>%
    ggplot(mapping = aes(x = daysabs, fill = prog)) +
    geom_bar(stat = "count") +
    facet_grid(prog ~ ., margin = TRUE, scales = "free_y") + 
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
          legend.key = element_blank(),
          plot.title = element_text(hjust = 0.5),
          strip.background = element_blank())
```

## Design matrix
I will assume the DAE design matrix is sufficient for modeling covariates.
```{r}
X <- model.matrix(object = daysabs ~ math + prog, data = data1)
head(X, n = 10)
```

## Poisson Model
$$\begin{align*}
  \eta_{i} &= \mathbf{X}_{i}^{T}\boldsymbol{\beta}\\
  \mu_{i} &= e^{\eta_{i}}\\
  y_{i} &\overset{\text{ind}}{\sim} Poisson(\mu_{i})\\
  \end{align*}
$$


--------------------
- Top Page: http://rpubs.com/kaz_yos/
- Github: https://github.com/kaz-yos

```{r}
print(sessionInfo())
## Record execution time and multicore use
end_time <- Sys.time()
diff_time <- difftime(end_time, start_time, units = "auto")
cat("Started  ", as.character(start_time), "\n",
    "Finished ", as.character(end_time), "\n",
    "Time difference of ", diff_time, " ", attr(diff_time, "units"), "\n",
    "Used ", foreach::getDoParWorkers(), " cores\n",
    "Used ", foreach::getDoParName(), " as backend\n",
    sep = "")
```
