---
title: "Dirichlet Process Mixture"
author: "Kazuki Yoshida"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: html_document
---

```{r, message = FALSE, tidy = FALSE, echo = F}
## knitr configuration: http://yihui.name/knitr/options#chunk_options
library(knitr)
showMessage <- FALSE
showWarning <- TRUE
set_alias(w = "fig.width", h = "fig.height", res = "results")
opts_chunk$set(comment = "##", error= TRUE, warning = showWarning, message = showMessage,
               tidy = FALSE, cache = F, echo = T,
               fig.width = 7, fig.height = 7, dev.args = list(family = "sans"))
## for rgl
## knit_hooks$set(rgl = hook_rgl, webgl = hook_webgl)
## for animation
opts_knit$set(animation.fun = hook_ffmpeg_html)
## R configuration
options(width = 116, scipen = 5)
## Record start time
start_time <- Sys.time()
## Configure parallelization
## Parallel backend for foreach (also loads foreach and parallel; includes doMC)
library(doParallel)
## Reproducible parallelization
library(doRNG)
## Detect core count (Do not use on clusters)
n_cores <- parallel::detectCores()
## Used by parallel::mclapply() as default
options(mc.cores = n_cores)
## Used by doParallel as default
options(cores = n_cores)
## Register doParallel as the parallel backend for foreach
## http://stackoverflow.com/questions/28989855/the-difference-between-domc-and-doparallel-in-r
doParallel::registerDoParallel(cores = n_cores)
```

## References
- Books
  - [(BNPDA) Bayesian Nonparametric Data Analysis](https://web.ma.utexas.edu/users/pmueller/bnp/)
  - [(HSAUR) A Handbook of Statistical Analyses using R](https://www.crcpress.com/A-Handbook-of-Statistical-Analyses-using-R/Hothorn-Everitt/p/book/9781482204582)
  - [(BIDA) Bayesian Ideas and Data Analysis An Introduction for Scientists and Statisticians](http://blogs.oregonstate.edu/bida/)
  - [(FNBI) Fundamentals of Nonparametric Bayesian Inference](https://www.cambridge.org/core/books/fundamentals-of-nonparametric-bayesian-inference/C96325101025D308C9F31F4470DEA2E8)
- Software
  - [BNPDA code](https://web.ma.utexas.edu/users/pmueller/bnp/)
  - [CRAN DPpackage: Bayesian Nonparametric Modeling in R](https://cran.r-project.org/package=DPpackage)
  - [J Stat Softw. 2011. DPpackage: Bayesian Non- and Semi-parametric Modelling in R.](https://www.jstatsoft.org/article/view/v040i05)
- Papers
  - [Chan et al (2017). Nonparametric estimation in economics: Bayesian and frequentist approaches.](https://onlinelibrary.wiley.com/doi/pdf/10.1002/wics.1406) (pre-print available)


## Load packages

```{r}
library(tidyverse)
library(DPpackage)
## library(MCMCpack)
```

## Dirichlet Process and Dirichlet Process Mixture

### Why Construct a Mixture
The [previous example of the Dirichlet Process](http://rpubs.com/kaz_yos/dp1) was based on discrete data with a discrete centering measure $G_{0}$. So it was not a surprise that we got a posterior distribution of discrete measures $G$. However, the Dirichlet process results in measures $G$ that are discrete almost surely (with probability one) regardless of the centering measure $G_{0}$. Therefore, for density estimation, adding one more layer to the model to use the DP as a mixture distribution is more appealing.

### Hierarchical Structure

$$\begin{align*}
  y_{i} | \theta_{i} &\overset{\text{ind}}{\sim} f_{\theta_{i}}\\
  \\
  \theta_{i} | G &\overset{\text{iid}}{\sim} G\\
  \\
  G &\sim DP(M G_{0})
  \end{align*}
$$

This is interpreted as follows.

1. Each data point $y_{i}$ has its own latent parameter value $\theta_{i}$. Given this latent parameter value, $y_{i} | \theta_{i}$ follows a parametric distribution with a density function $f_{\theta_{i}}$.

2. In turn, each individual-specific latent parameter value $\theta_{i}$ given a random probability measure $G$ is an iid sample from the distribution defined by $G$.

3. The distribution over the random measure $G$ is governed by a Dirichlet process with a mass (concentration) parameter $M$ and a centering measure $G_{0}$, where $G_{0}$ should have a support appropriate for $\theta_{i}$.

### Posterior Inference

Give the structure above and vector notations $\boldsymbol{\theta} = (\theta_{1},\dots,\theta_{n})$ and $\mathbf{y} = (y_{1}, \dots, y_{n})$. The posterior distribution of the random measure $G$ given $\boldsymbol{\theta}$ is the following.

$$\begin{align*}
  G | \boldsymbol{\theta} &\sim DP \left( MG_{0} + \sum^{n}_{i=1} \delta_{\theta_{i}} \right)
  \end{align*}
$$

Consider the posterior distribution of $\boldsymbol{\theta}$ given $\mathbf{y}$, $p(\boldsymbol{\theta} | \mathbf{y})$. If we integrate out $\boldsymbol{\theta}$ in the above expression using this posterior distribution, we obtain the following.

$$\begin{align*}
  G | \mathbf{y} &\sim \int DP \left( MG_{0} + \sum^{n}_{i=1} \delta_{\theta_{i}} \right) \text{d} p(\boldsymbol{\theta} | \mathbf{y})
  \end{align*}
$$

### Load data

This is an example from the BNPDA book (p13).

```{r}
## https://web.ma.utexas.edu/users/pmueller/bnp/R/DP/eig121/EIG.txt
data1 <- read.table(header = TRUE, text = "
  id recorded-EIG121 imputed-EIG121
  1      NA 0.00941
  2 0.00533 0.00533
  3 0.02310 0.02310
  4      NA 0.67871
  5 0.00008 0.00008
  6 0.27840 0.27840
  7 0.06530 0.06530
  8      NA 0.03108
  9      NA 0.03108
 10 0.00060 0.00060
 11 0.02750 0.02750
 12      NA 0.03108
 13 0.03060 0.03060
 14      NA 0.03108
 15      NA 0.67871
 16 0.03300 0.03300
 17      NA 0.03108
 18      NA 0.03108
 19      NA 0.67871
 20 0.09650 0.09650
 21 0.02560 0.02560
 22 0.01670 0.01670
 23 0.06980 0.06980
 24 0.00730 0.00730
 25      NA 0.03108
 26      NA 0.77071
 27      NA 0.03108
 28 0.00500 0.00500
 29      NA 0.03108
 31 0.00891 0.00891
 32      NA 0.22861
 33      NA 0.22861
 34 0.02630 0.02630
 35      NA 0.22861
 36      NA 0.03108
 37      NA 0.03108
 38      NA 0.03108
 39      NA 0.00941
 40 0.01357 0.01357
 41 0.01759 0.01759
 42 0.83840 0.83840
 43 1.18167 1.18167
 44 0.01540 0.01540
 45 0.67239 0.67239
 46 0.09752 0.09752
 47 0.16455 0.16455
 48 0.55606 0.55606
 49 0.64961 0.64961
 50 0.01287 0.01287
 51 0.00192 0.00192
 52 0.00043 0.00043
 53 0.01897 0.01897
 54 0.00036 0.00036
 55 0.23272 0.23272
 56 0.43327 0.43327
 57 1.10649 1.10649
 58 0.85073 0.85073
 59 0.00318 0.00318
 60 0.03394 0.03394
 61 0.13869 0.13869
 62 0.00345 0.00345
 63 0.22849 0.22849
 64 0.28294 0.28294
 65 0.75363 0.75363
 66 0.18504 0.18504
 67 0.01930 0.01930
 68 1.05398 1.05398
 69 0.02071 0.02071
 70 0.00295 0.00295
 71 0.00394 0.00394
 72 0.32358 0.32358
 73 0.01006 0.01006
 ")
```

--------------------
- Top Page: http://rpubs.com/kaz_yos/
- Github: https://github.com/kaz-yos

```{r}
print(sessionInfo())
## Record execution time and multicore use
end_time <- Sys.time()
diff_time <- difftime(end_time, start_time, units = "auto")
cat("Started  ", as.character(start_time), "\n",
    "Finished ", as.character(end_time), "\n",
    "Time difference of ", diff_time, " ", attr(diff_time, "units"), "\n",
    "Used ", foreach::getDoParWorkers(), " cores\n",
    "Used ", foreach::getDoParName(), " as backend\n",
    sep = "")
```
