---
title: "DPpackage"
author: "Kazuki Yoshida"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: html_document
---

```{r, message = FALSE, tidy = FALSE, echo = F}
## knitr configuration: http://yihui.name/knitr/options#chunk_options
library(knitr)
showMessage <- FALSE
showWarning <- TRUE
set_alias(w = "fig.width", h = "fig.height", res = "results")
opts_chunk$set(comment = "##", error= TRUE, warning = showWarning, message = showMessage,
               tidy = FALSE, cache = F, echo = T,
               fig.width = 7, fig.height = 7, dev.args = list(family = "sans"))
## for rgl
## knit_hooks$set(rgl = hook_rgl, webgl = hook_webgl)
## for animation
opts_knit$set(animation.fun = hook_ffmpeg_html)
## R configuration
options(width = 116, scipen = 5)
## Record start time
start_time <- Sys.time()
## Configure parallelization
## Parallel backend for foreach (also loads foreach and parallel; includes doMC)
library(doParallel)
## Reproducible parallelization
library(doRNG)
## Detect core count (Do not use on clusters)
n_cores <- parallel::detectCores()
## Used by parallel::mclapply() as default
options(mc.cores = n_cores)
## Used by doParallel as default
options(cores = n_cores)
## Register doParallel as the parallel backend for foreach
## http://stackoverflow.com/questions/28989855/the-difference-between-domc-and-doparallel-in-r
doParallel::registerDoParallel(cores = n_cores)
```

## References
- Books
  - [(BNPDA) Bayesian Nonparametric Data Analysis](https://www.springer.com/us/book/9783319189673)
- Software
  - [CRAN DPpackage: Bayesian Nonparametric Modeling in R](https://cran.r-project.org/package=DPpackage)
  - [J Stat Softw. 2011. DPpackage: Bayesian Non- and Semi-parametric Modelling in R.](https://www.jstatsoft.org/article/view/v040i05)
- Papers
  - [Guindani (2014). A Bayesian Semi-parametric Approach for the Differential Analysis of Sequence Counts Data.](https://www.ncbi.nlm.nih.gov/pubmed/24833809)


## Load packages

```{r}
library(tidyverse)
library(DPpackage)
```

## Dirichlet Process
Here we will use the T cell receptor example in BNPDA (p10) for demonstration of Dirichlet process prior.

### Data
The data is based on Guindani 2014. Note count 0 does not exist in the data by design. Counts equal to and greater than 5 were not observed.

```{r}
tcell <- tribble(
    ~count, ~frequency,
    1, 37,
    2, 11,
    3, 5,
    4, 2,
    ## >= 5, 0,
    )
tcell
```

### Model
The model is the following.

$$
y_{i} | G \overset{iid}\sim G\\
G \sim DP(M, G_{0})
$$

The notation $y_{i} | G \overset{iid}\sim G$ puzzled me for a while, but the LHS data | parameter is necessary because we consider the joint distribution of data and parameter in Bayesian statistics. An IID argument like seen in Frequentist paradigm is only possible after fixing (conditioning on) the parameter. Importantly, once we marginalize this conditioning over the prior distribution, exchangeability remains, but not generally IID.

In the Dirichlet process formulation, each random probability measure $G$ retains the same support as $G_{0}$. Here we will choose a Poisson with mean 2 that truncates at 1 (zeros are removed).


## Dirichlet Process Mixture


--------------------
- Top Page: http://rpubs.com/kaz_yos/
- Github: https://github.com/kaz-yos

```{r}
print(sessionInfo())
## Record execution time and multicore use
end_time <- Sys.time()
diff_time <- difftime(end_time, start_time, units = "auto")
cat("Started  ", as.character(start_time), "\n",
    "Finished ", as.character(end_time), "\n",
    "Time difference of ", diff_time, " ", attr(diff_time, "units"), "\n",
    "Used ", foreach::getDoParWorkers(), " cores\n",
    "Used ", foreach::getDoParName(), " as backend\n",
    sep = "")
```
