---
title: "Inverse-Gamma and Gamma distributions"
author: "Kazuki Yoshida"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: html_document
---

```{r, message = FALSE, tidy = FALSE, echo = F}
## knitr configuration: http://yihui.name/knitr/options#chunk_options
library(knitr)
showMessage <- FALSE
showWarning <- TRUE
set_alias(w = "fig.width", h = "fig.height", res = "results")
opts_chunk$set(comment = "##", error= TRUE, warning = showWarning, message = showMessage,
               tidy = FALSE, cache = F, echo = T,
               fig.width = 7, fig.height = 7, dev.args = list(family = "sans"))
## for rgl
## knit_hooks$set(rgl = hook_rgl, webgl = hook_webgl)
## for animation
opts_knit$set(animation.fun = hook_ffmpeg_html)
## R configuration
options(width = 116, scipen = 5)
## Record start time
start_time <- Sys.time()
## Configure parallelization
## Parallel backend for foreach (also loads foreach and parallel; includes doMC)
library(doParallel)
## Reproducible parallelization
library(doRNG)
## Detect core count (Do not use on clusters)
n_cores <- parallel::detectCores()
## Used by parallel::mclapply() as default
options(mc.cores = n_cores)
## Used by doParallel as default
options(cores = n_cores)
## Register doParallel as the parallel backend for foreach
## http://stackoverflow.com/questions/28989855/the-difference-between-domc-and-doparallel-in-r
doParallel::registerDoParallel(cores = n_cores)
```

## References
- [CRAN invgamma package](https://cran.r-project.org/web/packages/invgamma/invgamma.pdf)
- [Wikipedia Gamma distribution](https://en.wikipedia.org/wiki/Gamma_distribution)
- [Wikipedia Inverse-gamma distribution](https://en.wikipedia.org/wiki/Inverse-gamma_distribution)
- [Transformations of Random Variables](http://internet.math.arizona.edu/~jwatkins/f-transform.pdf)


## Load Packages

```{r}
library(tidyverse)
library(invgamma)
```

## Definitions

### Gamma Density

Let $\alpha > 0$ be the shape parameter and $\beta > 0$ be the rate parameter. For $x > 0$, the gamma density is the following.

$$\begin{align*}
  Gamma(x | \alpha, \beta) &= \frac{\beta^{\alpha}}{\Gamma(\alpha)} x^{\alpha - 1} e^{-\beta x}\\
  E(X) &= \frac{\alpha}{\beta}\\
  Var(X) &= \frac{\alpha}{\beta^2}\\
  \end{align*}
$$

### Inverse-Gamma Density

Let $\alpha > 0$ be the shape parameter and $\beta > 0$ be the rate parameter. For $x > 0$, the inverse-gamma density is the following.

$$\begin{align*}
  Inv-Gamma(x | \alpha, \beta) &= \frac{\beta^{\alpha}}{\Gamma(\alpha)} \left( \frac{1}{x} \right) ^{\alpha + 1} e^{-\beta \left( \frac{1}{x} \right)}\\
  E(X) &= \frac{\beta}{\alpha - 1} \text{ for } \alpha > 1\\
  Var(X) &= \frac{\beta^2}{(\alpha - 1)^2(\alpha - 2)} \text{ for } \alpha > 2\\
  \end{align*}
$$

### Relationship

Let $X$ be a random variable that follows $Gamma(\alpha,\beta)$. We want the distribution of $Y = g(X) = \frac{1}{X}$. Note the inverse transformation is $g^{-1}(y) = \frac{1}{y}$ and its derivative is $\frac{\text{d}}{\text{d}y} g^{-1}(y) = -\frac{1}{y^{2}}$.

$$\begin{align*}
  f_{Y}(y) &= f_{X}(g^{-1}(y)) \left| \frac{\text{d}}{\text{d}y} g^{-1}(y)\right|\\
  &= Gamma \left( \frac{1}{y} \bigg| \alpha, \beta \right) \left| -\frac{1}{y^{2}} \right|\\
  &= Gamma \left( \frac{1}{y} \bigg| \alpha, \beta \right) \frac{1}{y^{2}}\\
  &= \frac{\beta^{\alpha}}{\Gamma(\alpha)} \left( \frac{1}{y} \right)^{\alpha - 1} e^{-\beta \left( \frac{1}{y} \right)} \frac{1}{y^{2}}\\
  &= \frac{\beta^{\alpha}}{\Gamma(\alpha)} \left( \frac{1}{y} \right)^{\alpha + 1} e^{-\beta \left( \frac{1}{y} \right)}\\
  &= Inv-Gamma(y | \alpha, \beta)\\
  \end{align*}
$$

That is, when $X \sim Gamma(\alpha,\beta)$, $Y = \frac{1}{X} \sim Inv-Gamma(\alpha,\beta)$ with the same shape parameter $\alpha$ and rate parameter $\beta$.

## Visual Comparison
```{r}
plot(partial(dgamma, shape = 0.001, rate = 0.001))
plot(partial(dinvgamma, shape = 0.001, rate = 0.001))
```

```{r}
summary(rgamma(n = 10^4, shape = 0.001, rate = 0.001))
summary(1/rgamma(n = 10^4, shape = 0.001, rate = 0.001))
summary(rinvgamma(n = 10^4, shape = 0.001, rate = 0.001))
summary(rinvgamma(n = 10^4, shape = 0.01, rate = 0.01))
summary(rinvgamma(n = 10^4, shape = 0.1, rate = 0.1))
```

--------------------
- Top Page: http://rpubs.com/kaz_yos/
- Github: https://github.com/kaz-yos

```{r}
print(sessionInfo())
## Record execution time and multicore use
end_time <- Sys.time()
diff_time <- difftime(end_time, start_time, units = "auto")
cat("Started  ", as.character(start_time), "\n",
    "Finished ", as.character(end_time), "\n",
    "Time difference of ", diff_time, " ", attr(diff_time, "units"), "\n",
    "Used ", foreach::getDoParWorkers(), " cores\n",
    "Used ", foreach::getDoParName(), " as backend\n",
    sep = "")
```
