---
title: "BUGS Book Measurement Error"
author: "Kazuki Yoshida"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: html_document
---

```{r, message = FALSE, tidy = FALSE, echo = F}
## knitr configuration: http://yihui.name/knitr/options#chunk_options
library(knitr)
showMessage <- FALSE
showWarning <- TRUE
set_alias(w = "fig.width", h = "fig.height", res = "results")
opts_chunk$set(comment = "##", error= TRUE, warning = showWarning, message = showMessage,
               tidy = FALSE, cache = FALSE, echo = TRUE,
               fig.width = 7, fig.height = 7, dev.args = list(family = "sans"))
## for rgl
## knit_hooks$set(rgl = hook_rgl, webgl = hook_webgl)
## for animation
opts_knit$set(animation.fun = hook_ffmpeg_html)
## R configuration
options(width = 116, scipen = 5)
## Record start time
start_time <- Sys.time()
## Configure parallelization
## Parallel backend for foreach (also loads foreach and parallel; includes doMC)
library(doParallel)
## Reproducible parallelization
library(doRNG)
## Detect core count (Do not use on clusters)
n_cores <- parallel::detectCores()
## Used by parallel::mclapply() as default
options(mc.cores = n_cores)
## Used by doParallel as default
options(cores = n_cores)
## Register doParallel as the parallel backend for foreach
## http://stackoverflow.com/questions/28989855/the-difference-between-domc-and-doparallel-in-r
doParallel::registerDoParallel(cores = n_cores)
```

## References
- BUGS Book 9.3 Measurement Error
- BUGS Book Web Version http://www.openbugs.net/Examples/Cervix.html
- Stan Translation (Seems incorrect?) https://github.com/stan-dev/example-models/tree/master/bugs_examples/vol2/cervix
- Lambert. A Student's Guide to Bayesian Statistics. Chapter 16. p401.
- Gustafson. Measurement Error and Misclassification. Chapter 4. p55.

## Load packages

```{r}
library(tidyverse)
library(ggdag)
library(biostan)
library(rstan)
```

## Load data

The dataset is a case-control study on the outcome of invasive cervical cancer and the exposure of herpes simplex virus (HSV). Let $Y$ be the binary outcome, $X_{true}$ be the true binary covariate, and $X_{mis}$ be the mismeasured binary covariate. The dataset contains an internal validation dataset with both $X_{true}$ and $X_{mis}$. The rest of the dataset only has $X_{mis}$. Let $R_{X}$ be the indicator to indicate that $X_{true}$ was observed for that row.

```{r}
cervical <- tribble(
    ~Y, ~X_true, ~X_mis, ~count, ~R_X,
    ## Complete data for the internal validation set
    1, 0, 0, 13, 1,
    1, 0, 1,  3, 1,
    1, 1, 0,  5, 1,
    1, 1, 1, 18, 1,
    0, 0, 0, 33, 1,
    0, 0, 1, 11, 1,
    0, 1, 0, 16, 1,
    0, 1, 1, 16, 1,
    ## Incomplete data for the rest of the dataset
    1, NA, 0, 318, 0,
    1, NA, 1, 375, 0,
    0, NA, 0, 701, 0,
    0, NA, 1, 535, 0
)
cervical
## Stan does not take NAs.
cervical <- cervical %>%
    mutate(X_true = if_else(is.na(X_true), -1, X_true))
```

## Model

The model considered is a classical mismeasurement model represented in the following model DAG that also contains parameters. This type of DAGs with parameters in them are commonly seen in the Bayesian literature, in particular to describe hierarchical models. Here we also have the missing indicator assuming MCAR (validation subset is a random subsample within the sample).

```{r}
dagify(Y ~ Xt, Xm ~ Xt, Xm ~ phi, Xt ~ psi, Y ~ beta, Rx ~ gamma) %>%
    ggdag()
```

Since the missing process is ignorable, we can work on the other parts only. We partition the true exposure $X_{true}$ into the complete part $X_{true}^{c}$ and the reduced part $X_{true}^{r}$ (missing part). Let $i=1,\dots,m$ index the reduced part and $i=m+1,\dots,n$ index the complete part.

$$\begin{align*}
  p(X_{true}^{r}, \beta, \phi, \psi | Y, X_{mis}, X_{true}^{c})
  &\propto p(\beta, \phi, \psi, Y, X_{mis}, X_{true})\\
  \\
  &=
    \underbrace{p(\beta, \phi, \psi)}_{\text{prior}}
    \underbrace{p(Y, X_{mis}, X_{true} | \beta, \phi, \psi)}_{\text{likelihood}}\\
  \\
  &~~~~\text{By iid given parameters}\\
  &= p(\beta, \phi, \psi)
    \prod^{n}_{i=1} p(Y_{i}, X_{mis,i}, X_{true,i} | \beta, \phi, \psi)\\
  \\
  &~~~~\text{Partition again}\\
  &= p(\beta, \phi, \psi)\\
  &~~~\times \prod^{m}_{i=1}   p(Y_{i}, X_{mis,i}^{r}, X_{true,i} | \beta, \phi, \psi)\\
  &~~~\times \prod^{n}_{j=m+1} p(Y_{j}, X_{mis,j}^{c}, X_{true,j} | \beta, \phi, \psi)\\
  \\
  &~~~~\text{Factor likelihood following DAG}\\
  &= p(\beta, \phi, \psi)\\
  &~~~\times \prod^{m}_{i=1}
    \underbrace{p(Y_{i} | X_{true,i}^{r}, \beta)}_{\text{outcome model}}
    \underbrace{p(X_{mis,i} | X_{true,i}^{r}, \phi)}_{\text{error model}}
    \underbrace{p(X_{true,i}^{r} | \psi)}_{\text{covariate model}}\\
  &~~~\times \prod^{n}_{j=m+1}
    \underbrace{p(Y_{j} | X_{true,j}^{c}, \beta)}_{\text{outcome model}}
    \underbrace{p(X_{mis,j} | X_{true,j}^{c}, \phi)}_{\text{error model}}
    \underbrace{p(X_{true,j}^{c} | \psi)}_{\text{covariate model}}\\
  \end{align*}
$$

Since we do not observe $X_{true}^{r}$, we need integrate out (really sum out here) this part from the joint posterior. That is,

$$\begin{align*}
  p(\beta, \phi, \psi | Y, X_{mis}, X_{true}^{c})
  &= \int p(X_{true}^{r}=x_{true}^{r}, \beta, \phi, \psi | Y, X_{mis}, X_{true}^{c}) dF(x_{true}^{r})
  \end{align*}
$$

The only part that depends on $X_{true}^{r}$ in the prior $\times$ likelihood expression is the following likelihood for the reduced part.

$$\begin{align*}
  \prod^{m}_{i=1}
    \underbrace{p(Y_{i} | X_{true,i}^{r}, \beta)}_{\text{outcome model}}
    \underbrace{p(X_{mis,i} | X_{true,i}^{r}, \phi)}_{\text{error model}}
    \underbrace{p(X_{true,i}^{r} | \psi)}_{\text{covariate model}}\\
  \end{align*}
$$

Conditioning on the parameters, $X_{true,i}^{r}$ are independent from each other. Thus, we can sum out each term in the product.

$$\begin{align*}
  &\prod^{m}_{i=1}
  \sum^{1}_{x=0}
    p(Y_{i} | X_{true,i}^{r}=x, \beta)
    p(X_{mis,i} | X_{true,i}^{r}=x, \phi)
    p(X_{true,i}^{r}=x | \psi)\\
  &= \prod^{m}_{i=1}
    \left[
    p(Y_{i} | X_{true,i}^{r}=0, \beta)
    p(X_{mis,i} | X_{true,i}^{r}=0, \phi)
    p(X_{true,i}^{r}=0 | \psi) \right.\\
  &~~~~~~~~~~~\left. + p(Y_{i} | X_{true,i}^{r}=1, \beta)
    p(X_{mis,i} | X_{true,i}^{r}=1, \phi)
    p(X_{true,i}^{r}=1 | \psi) \right]\\
  \\
  &= \prod^{m}_{i=1}
    \left[
    p(Y_{i} | X_{true,i}^{r}=0, \beta)
    p(X_{mis,i} | X_{true,i}^{r}=0, \phi)
    (1-\psi) \right.\\
  &~~~~~~~~~~~\left. + p(Y_{i} | X_{true,i}^{r}=1, \beta)
    p(X_{mis,i} | X_{true,i}^{r}=1, \phi)
    (\psi) \right]\\
  \end{align*}
$$

Therefore, the marginalized posterior and the corresponding prior $\times$ likelihood expression are the following.

$$\begin{align*}
  p(\beta, \phi, \psi | Y, X_{mis}, X_{true}^{c})
  &= \int p(X_{true}^{r}=x_{true}^{r}, \beta, \phi, \psi | Y, X_{mis}, X_{true}^{c}) dF(x_{true}^{r})\\
  \\
  &\propto p(\beta, \phi, \psi)\\
  &~~~\times \prod^{m}_{i=1}
    \left[
    p(Y_{i} | X_{true,i}^{r}=0, \beta)
    p(X_{mis,i} | X_{true,i}^{r}=0, \phi)
    (1-\psi) \right.\\
  &~~~~~~~~~~~~~~\left. + p(Y_{i} | X_{true,i}^{r}=1, \beta)
    p(X_{mis,i} | X_{true,i}^{r}=1, \phi)
    (\psi) \right]\\
  &~~~\times \prod^{n}_{j=m+1}
    p(Y_{j} | X_{true,j}^{c}, \beta)
    p(X_{mis,j} | X_{true,j}^{c}, \phi)
    p(X_{true,j}^{c} | \psi)\\
  \end{align*}
$$

The corresponding log expression is the following.

$$\begin{align*}
  \log p(\beta, \phi, \psi | Y, X_{mis}, X_{true}^{c})
  &\propto \log p(\beta, \phi, \psi)\\
  &~~~+ \sum^{m}_{i=1}\log
    \left[
    p(Y_{i} | X_{true,i}^{r}=0, \beta)
    p(X_{mis,i} | X_{true,i}^{r}=0, \phi)
    (1-\psi) \right.\\
  &~~~~~~~~~~~~~~~~~\left. + p(Y_{i} | X_{true,i}^{r}=1, \beta)
    p(X_{mis,i} | X_{true,i}^{r}=1, \phi)
    (\psi) \right]\\
  &~~~+ \sum^{n}_{j=m+1} \left[
    \log p(Y_{j} | X_{true,j}^{c}, \beta)+
    \log p(X_{mis,j} | X_{true,j}^{c}, \phi)+
    \log p(X_{true,j}^{c} | \psi) \right]\\
  \end{align*}
$$

Each term in the first summation can be written safely with the log_sum_exp() function in Stan.

$$\begin{align*}
  &\log
    \left[
    p(Y_{i} | X_{true,i}^{r}=0, \beta)
    p(X_{mis,i} | X_{true,i}^{r}=0, \phi)
    (1-\psi) \right.\\
  &~~~~~~\left. + p(Y_{i} | X_{true,i}^{r}=1, \beta)
    p(X_{mis,i} | X_{true,i}^{r}=1, \phi)
    (\psi) \right]\\
  &= \text{log_sum_exp}
    \left( \right.\\
  &~~~~~~\log p(Y_{i} | X_{true,i}^{r}=0, \beta) +
    \log p(X_{mis,i} | X_{true,i}^{r}=0, \phi) +
    \log (1-\psi),\\
  &~~~~~~\log p(Y_{i} | X_{true,i}^{r}=1, \beta) +
    \log p(X_{mis,i} | X_{true,i}^{r}=1, \phi) +
    \log (\psi)\\
  &~~~\left. \right)\\
  \end{align*}
$$

For an individual without the gold standard measurement, $p(X_{true,i}^{r} = 1 | \psi) = \psi$ before data. Conditioning on the observed data ($X_{mis,i}$ and $Y_{i}$) and parameters, the distribution of the unobserved $X_{true,i}$ is the following.

$$\begin{align*}
  p(X_{true,i}^{r} = 1 | X_{mis,i}, Y_{i}, \beta, \phi, \psi)
  &= \frac{p(X_{true,i}^{r} = 1, X_{mis,i}, Y_{i}, \beta, \phi, \psi)}{\sum^{1}_{x=0} p(X_{true,i}^{r} = 1, X_{mis,i}, Y_{i}, \beta, \phi, \psi)}\\
  &~~~~\text{Using $q$ for unnormalized density}\\
  &= \frac{q(X_{true,i}^{r} = 1, X_{mis,i}, Y_{i}, \beta, \phi, \psi)}{\sum^{1}_{x=0} q(X_{true,i}^{r} = 1, X_{mis,i}, Y_{i}, \beta, \phi, \psi)}\\
  \\
  &~~~~\text{Also}\\
  \log p(X_{true,i}^{r} = 1 | X_{mis,i}, Y_{i}, \beta, \phi, \psi)
  &= \log q(X_{true,i}^{r} = 1, X_{mis,i}, Y_{i}, \beta, \phi, \psi) - \log \left[ \sum^{1}_{x=0} q(X_{true,i}^{r} = 1, X_{mis,i}, Y_{i}, \beta, \phi, \psi) \right]\\
  \end{align*}
$$

Thus, we can reuse the log_sum_exp values for normalization to obtain $p(X_{true,i}^{r} = 1 | X_{mis,i}, Y_{i}, \beta, \phi, \psi)$.

## Stan implementation

```{r}
model1_code <- biostan::read_stan_file("./mismeasurement_bugs_book.stan")
biostan::print_stan_code(model1_code)
```
```{r, results = "hide", cache = FALSE}
model1 <- stan(model_code = model1_code,
               data = c(N = nrow(cervical),
                        as.list(cervical)),
               cores = n_cores)
```
```{r}
model1
traceplot(model1, pars = names(model1))
pairs(model1)
```

The convergence looks ok. The results are consistent with the example fit in BUGS.

--------------------
- Top Page: http://rpubs.com/kaz_yos/
- Github: https://github.com/kaz-yos

```{r}
print(sessionInfo())
## Record execution time and multicore use
end_time <- Sys.time()
diff_time <- difftime(end_time, start_time, units = "auto")
cat("Started  ", as.character(start_time), "\n",
    "Finished ", as.character(end_time), "\n",
    "Time difference of ", diff_time, " ", attr(diff_time, "units"), "\n",
    "Used ", foreach::getDoParWorkers(), " cores\n",
    "Used ", foreach::getDoParName(), " as backend\n",
    sep = "")
```
