---
title: "Polya Tree Prior"
author: "Kazuki Yoshida"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: html_document
---

```{r, message = FALSE, tidy = FALSE, echo = F}
## knitr configuration: http://yihui.name/knitr/options#chunk_options
library(knitr)
showMessage <- FALSE
showWarning <- TRUE
set_alias(w = "fig.width", h = "fig.height", res = "results")
opts_chunk$set(comment = "##", error= TRUE, warning = showWarning, message = showMessage,
               tidy = FALSE, cache = F, echo = T,
               fig.width = 7, fig.height = 7, dev.args = list(family = "sans"))
## for rgl
## knit_hooks$set(rgl = hook_rgl, webgl = hook_webgl)
## for animation
opts_knit$set(animation.fun = hook_ffmpeg_html)
## R configuration
options(width = 116, scipen = 5)
## Record start time
start_time <- Sys.time()
## Configure parallelization
## Parallel backend for foreach (also loads foreach and parallel; includes doMC)
library(doParallel)
## Reproducible parallelization
library(doRNG)
## Detect core count (Do not use on clusters)
n_cores <- parallel::detectCores()
## Used by parallel::mclapply() as default
options(mc.cores = n_cores)
## Used by doParallel as default
options(cores = n_cores)
## Register doParallel as the parallel backend for foreach
## http://stackoverflow.com/questions/28989855/the-difference-between-domc-and-doparallel-in-r
doParallel::registerDoParallel(cores = n_cores)
```

## References
- Books
  - [(BNPDA) Bayesian Nonparametric Data Analysis](https://www.springer.com/us/book/9783319189673)
  - [(FNBI) Fundamentals of Nonparametric Bayesian Inference](https://www.cambridge.org/core/books/fundamentals-of-nonparametric-bayesian-inference/C96325101025D308C9F31F4470DEA2E8)
- Software
  - [BNPDA code](https://web.ma.utexas.edu/users/pmueller/bnp/)
  - [CRAN DPpackage: Bayesian Nonparametric Modeling in R](https://cran.r-project.org/package=DPpackage)
  - [J Stat Softw. 2011. DPpackage: Bayesian Non- and Semi-parametric Modelling in R.](https://www.jstatsoft.org/article/view/v040i05)
- Papers
  - [Guindani (2014). A Bayesian Semi-parametric Approach for the Differential Analysis of Sequence Counts Data.](https://www.ncbi.nlm.nih.gov/pubmed/24833809)


## Load packages

```{r}
library(tidyverse)
library(DPpackage)
library(MCMCpack)
```

## Polya Tree




--------------------
- Top Page: http://rpubs.com/kaz_yos/
- Github: https://github.com/kaz-yos

```{r}
print(sessionInfo())
## Record execution time and multicore use
end_time <- Sys.time()
diff_time <- difftime(end_time, start_time, units = "auto")
cat("Started  ", as.character(start_time), "\n",
    "Finished ", as.character(end_time), "\n",
    "Time difference of ", diff_time, " ", attr(diff_time, "units"), "\n",
    "Used ", foreach::getDoParWorkers(), " cores\n",
    "Used ", foreach::getDoParName(), " as backend\n",
    sep = "")
```
