---
title: "SAS PROC MIXED in Stan"
author: "Kazuki Yoshida"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: html_document
---

```{r, message = FALSE, tidy = FALSE, echo = F}
## knitr configuration: http://yihui.name/knitr/options#chunk_options
library(knitr)
showMessage <- FALSE
showWarning <- TRUE
set_alias(w = "fig.width", h = "fig.height", res = "results")
opts_chunk$set(comment = "##", error= TRUE, warning = showWarning, message = showMessage,
               tidy = FALSE, cache = FALSE, echo = TRUE,
               fig.width = 7, fig.height = 7, dev.args = list(family = "sans"))
## for rgl
## knit_hooks$set(rgl = hook_rgl, webgl = hook_webgl)
## for animation
opts_knit$set(animation.fun = hook_ffmpeg_html)
## R configuration
options(width = 116, scipen = 5)
## Record start time
start_time <- Sys.time()
## Configure parallelization
## Parallel backend for foreach (also loads foreach and parallel; includes doMC)
library(doParallel)
## Reproducible parallelization
library(doRNG)
## Detect core count (Do not use on clusters)
n_cores <- parallel::detectCores()
## Used by parallel::mclapply() as default
options(mc.cores = n_cores)
## Used by doParallel as default
options(cores = n_cores)
## Register doParallel as the parallel backend for foreach
## http://stackoverflow.com/questions/28989855/the-difference-between-domc-and-doparallel-in-r
doParallel::registerDoParallel(cores = n_cores)
```

## Introduction
SAS PROC MIXED is a powerful procedure for Gaussian data which conducts linear mixed models through the RANDOM statement. Using the RANDOM statement in isolation, the residuals conditional on the random effects are assumed to be independent within each individual, $\sigma^2\mathbf{I}$. Additionally, this procedure can model the covariance of residuals through the REPEATED statement. This can be used in isolation, which gives multivariate normal models, not mixed effects models. It can also be combined with the RANDOM statement to model the remaining dependent structure in the conditional residuals. Here we will examine what these mean by implementing comparable Bayesian models in Stan.

## References
- SAS papers
  - [Repeated Measures Modeling With PROC MIXED](https://support.sas.com/resources/papers/proceedings/proceedings/sugi29/188-29.pdf)
  - [RANDOM and REPEATED statements - How to Use Them to Model the Covariance Structure in Proc Mixed](https://www.mwsug.org/proceedings/2007/stats/MWSUG-2007-S02.pdf)
- SAS documentation
  - [Overview: MIXED Procedure](https://support.sas.com/documentation/cdl/en/statug/63033/HTML/default/viewer.htm#statug_mixed_sect001.htm)
  - [SAS PROC MIXED RANDOM statement](https://support.sas.com/documentation/cdl/en/statug/63033/HTML/default/viewer.htm#statug_mixed_sect018.htm)
  - [SAS PROC MIXED REPEATED statement](https://support.sas.com/documentation/cdl/en/statug/63033/HTML/default/viewer.htm#statug_mixed_sect019.htm)
  - [SAS Table 56.15 Covariance Structure Examples](https://support.sas.com/documentation/cdl/en/statug/63033/HTML/default/viewer.htm#statug_mixed_sect019.htm)
- Covariance modeling papers
  - [Wolfinger 1996. Heterogeneous Variance: Covariance Structures for Repeated Measures](https://www.jstor.org/stable/pdf/1400366.pdf?seq=1#page_scan_tab_contents)
  - [Littell et al. 2000. TUTORIAL IN BIOSTATISTICS Modelling covariance structure in the analysis of repeated measures data](https://faculty.washington.edu/heagerty/Courses/VA-longitudinal/private/Littell-StatMed2000.pdf)
- Others
  - [Pig weights data](http://matt-wand.utsacademics.info/webspr/pig.weights.html)
  - [(BIDA) Chap 10.3 of Bayesian Ideas and Data Analysis An Introduction for Scientists and Statisticians](http://blogs.oregonstate.edu/bida/)

## Load packages
```{r}
library(tidyverse)
library(rstan)
```

## Load data
Based on the reference "Repeated Measures Modeling With PROC MIXED", we will use the pig weight data. It is a balanced design measurements of weights over 8 weeks (9 measurements) on 48 pigs.
```
Pig weight data
Description:
     The pig.weights data frame has 9 repeated weight measures on 48
     pigs.
Format:
     This data frame contains the following columns:
     id.num:    identification number of pig.
     num.weeks: number of weeks since measurements commenced.
     weight:    bodyweight of pig "id.num" after "num.weeks" weeks.
Source:
     Diggle, P.J., Heagerty, P., Liang, K.-Y. and Zeger, S.L. (2002).
     _Analysis of Longitudinal Data, Second Edition_, Oxord: Oxford
     University Press.
```
```{r}
data(pig.weights, package = "SemiPar")
pig.weights <- as_tibble(pig.weights)
pig.weights %>%
    group_by(id.num) %>%
    summarize(cluster_size = n()) %>%
    group_by(cluster_size) %>%
    summarise(n_clusters = n())
```
The trajectories are nearly linear within each individual. As often seen in a longitudinal dataset, the variability across individual is greater at a later time point.
```{r}
ggplot(data = pig.weights, mapping = aes(x = num.weeks, y = weight, group = id.num)) +
    geom_line() +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
          legend.key = element_blank(),
          plot.title = element_text(hjust = 0.5),
          strip.background = element_blank())
```

This is numerically seen as follows.
```{r}
pig.weights %>%
    group_by(num.weeks) %>%
    summarize(mean = mean(weight),
              var = var(weight),
              sd = sd(weight))
```
We will use a matrix version for model fitting as it is a balanced design data.
```{r}
pig.weights_wide <- pig.weights %>%
    spread(key = id.num, value = weight)
pig.weights_mat <- as.matrix(pig.weights_wide[,-1])
dim(pig.weights_mat)
```

Let $Y_{ij}$ be the weight measurement on the i-th pig (i = 1, ..., I = 48) at j-th week (j = 1, ..., J = 9). For each individual i, we have a vector-valued outcome $\mathbf{Y}_{i}$ ($J\times1$). Across individuals these vector-valued outcomes are independent and identically distributed with a mean vector $\boldsymbol{\mu}$ and a covariance matrix $\boldsymbol{\Sigma}$. In the following, we will assume multivariate normality.

## Correlated Errors
### Unstructured covariance structure
```
Proc Mixed Data=Pigs Update;
    Class Week;
    Model Weight = Week / DDFM=KenwardRoger;
    Repeated Week / Subject=Pig Type=UN HLM R RCorr RI;
    LSMeans Week;
Run;
```
$$\begin{align*}
  \mathbf{Y}_{i}|(\boldsymbol{\mu},\boldsymbol{\Sigma}) \overset{\text{iid}}{\sim} MVN(\boldsymbol{\mu},\boldsymbol{\Sigma})
  \end{align*}
$$
In this model, we do not have any constraint on the mean vector because of the categorically specified Week variable. Thus, $\boldsymbol{\mu}$ has 9 parameters. Also the covariance matrix is unstructured with the only constraint being symmetry and positive definiteness. Thus, $\boldsymbol{\Sigma}$ has 9 diagonal (variance) and 36 off-diagonal (covariance) parameters.

```{r}
unstr_stan <- rstan::stan_model(file = "./proc_mixed_repeated_unstr.stan")
unstr_stan
```
```{r, results="hide"}
unstr_stan_fit <-
    rstan::sampling(object = unstr_stan,
                    data = list(I = 48,
                                J = 9,
                                Mu_means = rep(0,9),
                                Mu_sds = rep(100,9),
                                sigma_cauchy_scale = rep(100,9),
                                L_eta = 0.1,
                                weight = pig.weights_mat,
                                use_lik = 1),
                    chains = n_cores,
                    cores = n_cores)
```
```{r, results="hide"}
unstr_stan_fit2 <-
    rstan::sampling(object = unstr_stan,
                    data = list(I = 48,
                                J = 9,
                                Mu_means = rep(0,9),
                                Mu_sds = rep(100,9),
                                sigma_cauchy_scale = rep(100,9),
                                L_eta = 0.01,
                                weight = pig.weights_mat,
                                use_lik = 1),
                    chains = n_cores,
                    cores = n_cores)
```
```{r}
## Check HMC diagnostics after sampling
rstan::check_hmc_diagnostics(unstr_stan_fit)
## Specify relevant parameters
pars <- c("lp__","Mu","Sigma")
## Print a summary for a fitted model represented by a 'stanfit' object
print(unstr_stan_fit, pars = pars)
```
```{r, eval = FALSE}
## Create a matrix of output plots from a 'stanfit' object
pairs(unstr_stan_fit, pars = c("Mu"))
pairs(unstr_stan_fit, pars = c("sigma"))
## Markov chain traceplots
rstan::traceplot(unstr_stan_fit, pars = c("Mu"), inc_warmup = FALSE)
rstan::traceplot(unstr_stan_fit, pars = c("sigma"), inc_warmup = FALSE)
## Trace plots of MCMC draws
regex_pars <- c("Mu")
bayesplot::mcmc_rank_hist(unstr_stan_fit, regex_pars = regex_pars, ref_line = TRUE)
bayesplot::mcmc_rank_overlay(unstr_stan_fit, regex_pars = regex_pars, ref_line = TRUE)
```
```{r}
as.data.frame(unstr_stan_fit, pars = c("Sigma")) %>%
    as_tibble() %>%
    gather(key = key, value = value) %>%
    mutate(key = gsub("Sigma\\[|\\]","",key)) %>%
    separate(key, c("row","col"), sep = ",") %>%
    group_by(row,col) %>%
    summarize(mean = mean(value)) %>%
    spread(key = col, value = mean)
```


### Heterogeneous variance autoregressive (Order 1) covariance structure
```
Proc Mixed Data=Pigs Update;
    Class Week;
    Model Weight = Week / DDFM=KenwardRoger;
    Repeated Week / Subject=Pig Type=ARH(1) R RCorr RI;
    LSMeans Week;
Run;
```
$$\begin{align*}
  \mathbf{Y}_{i} &\overset{\text{iid}}{\sim} MVN(\boldsymbol{\mu},\boldsymbol{\Sigma}_{AR})
  \end{align*}
$$
The mean specification is the same. Now the variance covariance matrix $\boldsymbol{\Sigma}_{AR}$ is more structured. There are still 9 diagonal (variance) parameters. The correlation matrix, which determines the off-diagonals along with the variance parameters, has the AR(1) structure. That is, neighboring outcomes (distance 1) within an individual have a correlation of $\rho$. Outcomes with a distance of 2 have a correlation of $\rho^2$. It keeps decaying until $Corr(Y_{i1},Y_{i9}) = \rho^8$. Therefore, the covariance matrix has 10 parameters.

```{r}
arh1_stan <- rstan::stan_model(file = "./proc_mixed_repeated_arh1.stan")
arh1_stan
```
```{r, results="hide"}
arh1_stan_fit <-
    rstan::sampling(object = arh1_stan,
                    data = list(I = 48,
                                J = 9,
                                Mu_means = rep(0,9),
                                Mu_sds = rep(100,9),
                                sigma_cauchy_scale = rep(100,9),
                                rho_a = 1,
                                rho_b = 1,
                                weight = pig.weights_mat,
                                use_lik = 1),
                    chains = n_cores,
                    cores = n_cores)
```
```{r}
## Check HMC diagnostics after sampling
rstan::check_hmc_diagnostics(arh1_stan_fit)
## Specify relevant parameters
pars <- c("lp__","Mu","sigma","rho")
## Print a summary for a fitted model represented by a 'stanfit' object
print(arh1_stan_fit, pars = pars)
```
```{r, eval = FALSE}
## Create a matrix of output plots from a 'stanfit' object
pairs(arh1_stan_fit, pars = c("Mu"))
pairs(arh1_stan_fit, pars = c("sigma"))
## Markov chain traceplots
rstan::traceplot(arh1_stan_fit, pars = c("Mu"), inc_warmup = FALSE)
rstan::traceplot(arh1_stan_fit, pars = c("sigma"), inc_warmup = FALSE)
## Trace plots of MCMC draws
regex_pars <- c("Mu")
bayesplot::mcmc_rank_hist(arh1_stan_fit, regex_pars = regex_pars, ref_line = TRUE)
bayesplot::mcmc_rank_overlay(arh1_stan_fit, regex_pars = regex_pars, ref_line = TRUE)
```
The posterior mean of each element of the covariance matrix is the following.
```{r}
as.data.frame(arh1_stan_fit, pars = c("Sigma")) %>%
    as_tibble() %>%
    gather(key = key, value = value) %>%
    mutate(key = gsub("Sigma\\[|\\]","",key)) %>%
    separate(key, c("row","col"), sep = ",") %>%
    group_by(row,col) %>%
    summarize(mean = mean(value)) %>%
    spread(key = col, value = mean)
```

### Heterogeneous variance Toeplitz covariance structure
```
Proc Mixed Data=Pigs Update;
    Class Week;
    Model Weight = Week / DDFM=KenwardRoger;
    Repeated Week / Subject=Pig Type=TOEPH R RCorr RI;
    LSMeans Week;
Run;
```
$$\begin{align*}
  \mathbf{Y}_{i} &\overset{\text{iid}}{\sim} MVN(\boldsymbol{\mu},\boldsymbol{\Sigma}_{AR})
  \end{align*}
$$
The mean specification is the same. Now the variance covariance matrix $\boldsymbol{\Sigma}_{AR}$ is more structured. There are still 9 diagonal (variance) parameters. The correlation matrix, which determines the off-diagonals along with the variance parameters, has the AR(1) structure. That is, neighboring outcomes (distance 1) within an individual have a correlation of $\rho$. Outcomes with a distance of 2 have a correlation of $\rho^2$. It keeps decaying until $Corr(Y_{i1},Y_{i9}) = \rho^8$. Therefore, the covariance matrix has 10 parameters.

```{r}
toeph1_stan <- rstan::stan_model(file = "./proc_mixed_repeated_toeph1.stan")
toeph1_stan
```
```{r, results="hide"}
toeph1_stan_fit <-
    rstan::sampling(object = toeph1_stan,
                    data = list(I = 48,
                                J = 9,
                                Mu_means = rep(0,9),
                                Mu_sds = rep(100,9),
                                sigma_cauchy_scale = rep(100,9),
                                rho_a = rep(1,8),
                                rho_b = rep(1,8),
                                weight = pig.weights_mat,
                                use_lik = 1))
```
```{r}
## Check HMC diagnostics after sampling
rstan::check_hmc_diagnostics(toeph1_stan_fit)
## Specify relevant parameters
pars <- c("lp__","Mu","Sigma")
## Print a summary for a fitted model represented by a 'stanfit' object
print(toeph1_stan_fit, pars = pars)
```
```{r, eval = FALSE}
## Create a matrix of output plots from a 'stanfit' object
pairs(toeph1_stan_fit, pars = c("Mu"))
pairs(toeph1_stan_fit, pars = c("sigma"))
## Markov chain traceplots
rstan::traceplot(toeph1_stan_fit, pars = c("Mu"), inc_warmup = FALSE)
rstan::traceplot(toeph1_stan_fit, pars = c("sigma"), inc_warmup = FALSE)
## Trace plots of MCMC draws
regex_pars <- c("Mu")
bayesplot::mcmc_rank_hist(toeph1_stan_fit, regex_pars = regex_pars, ref_line = TRUE)
bayesplot::mcmc_rank_overlay(toeph1_stan_fit, regex_pars = regex_pars, ref_line = TRUE)
```

### Heterogeneous variance antedependence covariance structure
```
Proc Mixed Data=Pigs Update;
    Class Week;
    Model Weight = Week / DDFM=KenwardRoger;
    Repeated Week / Subject=Pig Type=ANTE(1) R RCorr RI;
    LSMeans Week;
Run;
```

## Random coefficients
###
```
Proc Sort Data=Pigs;
    By Pig Week;
Run;
   Proc Mixed Data=Pigs Update;
    Model Weight = Week / Solution OutPM=PAFit OutP=SSFit;
    Random Intercept Week / Solution Subject=Pig Type=UN G GCorr V VCorr;
    Estimate "Week 1 Ave Weight" Intercept 1 Week 1;
    Estimate "Week 9 Ave Weight" Intercept 1 Week 9;
Run;
```
--------------------
- Top Page: http://rpubs.com/kaz_yos/
- Github: https://github.com/kaz-yos

```{r}
print(sessionInfo())
## Record execution time and multicore use
end_time <- Sys.time()
diff_time <- difftime(end_time, start_time, units = "auto")
cat("Started  ", as.character(start_time), "\n",
    "Finished ", as.character(end_time), "\n",
    "Time difference of ", diff_time, " ", attr(diff_time, "units"), "\n",
    "Used ", foreach::getDoParWorkers(), " cores\n",
    "Used ", foreach::getDoParName(), " as backend\n",
    sep = "")
```
