---
title: "PROC MIXED in Stan"
author: "Kazuki Yoshida"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: html_document
---

```{r, message = FALSE, tidy = FALSE, echo = F}
## knitr configuration: http://yihui.name/knitr/options#chunk_options
library(knitr)
showMessage <- FALSE
showWarning <- TRUE
set_alias(w = "fig.width", h = "fig.height", res = "results")
opts_chunk$set(comment = "##", error= TRUE, warning = showWarning, message = showMessage,
               tidy = FALSE, cache = FALSE, echo = TRUE,
               fig.width = 7, fig.height = 7, dev.args = list(family = "sans"))
## for rgl
## knit_hooks$set(rgl = hook_rgl, webgl = hook_webgl)
## for animation
opts_knit$set(animation.fun = hook_ffmpeg_html)
## R configuration
options(width = 116, scipen = 5)
## Record start time
start_time <- Sys.time()
## Configure parallelization
## Parallel backend for foreach (also loads foreach and parallel; includes doMC)
library(doParallel)
## Reproducible parallelization
library(doRNG)
## Detect core count (Do not use on clusters)
n_cores <- parallel::detectCores()
## Used by parallel::mclapply() as default
options(mc.cores = n_cores)
## Used by doParallel as default
options(cores = n_cores)
## Register doParallel as the parallel backend for foreach
## http://stackoverflow.com/questions/28989855/the-difference-between-domc-and-doparallel-in-r
doParallel::registerDoParallel(cores = n_cores)
```

## References
- https://twitter.com/DrSamanthaSeals/status/1151143564094689281
- https://support.sas.com/documentation/cdl/en/statug/63033/HTML/default/viewer.htm#statug_mixed_sect019.htm
- https://support.sas.com/resources/papers/proceedings/proceedings/sugi29/188-29.pdf
- [Pig weights data](http://matt-wand.utsacademics.info/webspr/pig.weights.html)

## Load packages
```{r}
library(tidyverse)
library(rstan)
```

## Load data
Based on the "Repeated Measures Modeling With PROC MIXED", we will use the pig weight data. It is a balanced design measurements of weights over 8 weeks (9 measurements) on 48 pigs.
```{r}
data(pig.weights, package = "SemiPar")
pig.weights <- as_tibble(pig.weights)
pig.weights %>%
    group_by(id.num) %>%
    summarize(cluster_size = n()) %>%
    group_by(cluster_size) %>%
    summarise(n_clusters = n())
```
```{r}
ggplot(data = pig.weights, mapping = aes(x = num.weeks, y = weight, group = id.num)) +
    geom_line() +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
          legend.key = element_blank(),
          plot.title = element_text(hjust = 0.5),
          strip.background = element_blank())
```
We will use a matrix version as it is a balanced design data.
```{r}
pig.weights_wide <- pig.weights %>%
    spread(key = id.num, value = weight)
pig.weights_mat <- as.matrix(pig.weights_wide[,-1])
dim(pig.weights_mat)
```

Let $Y_{ij}$ be the weight measurement on the i-th pig (i = 1, ..., I = 48) at j-th week (j = 1, ..., J = 9). For each individual i, we have a vector-valued outcome $\mathbf{Y}_{i}$ ($J\times1$). Across individuals these vector-valued outcomes are independent and identically distributed with a mean vector $\boldsymbol{\mu}$ and a covariance matrix $\boldsymbol{\Sigma}$. In the following, we will assume multivariate normality.

## Correlated Errors
### Unstructured covariance structure
```
Proc Mixed Data=Pigs Update;
    Class Week;
    Model Weight = Week / DDFM=KenwardRoger;
    Repeated Week / Subject=Pig Type=UN HLM R RCorr RI;
    LSMeans Week;
Run;
```
$$\begin{align*}
  \mathbf{Y}_{i} \overset{\text{iid}}{\sim} MVN(\boldsymbol{\mu},\boldsymbol{\Sigma})
  \end{align*}
$$
In this model, we do not have any constraint on the mean vector because of the categorically specified Week variable. Thus, $\boldsymbol{\mu}$ has 9 free parameters. Also the covariance matrix is unstructured with the only constraint being symmetry and positive definiteness. Thus, $\boldsymbol{\Sigma}$ has 9 diagonal (variance) and 36 off-diagonal (covariance) free parameters.

```{r}
unstr_stan <- rstan::stan_model(file = "./proc_mixed_repeated_unstr.stan")
unstr_stan
```
```{r, results="hide"}
unstr_stan_fit <-
    rstan::sampling(object = unstr_stan,
                    data = list(I = 48,
                                J = 9,
                                Mu_means = rep(0,9),
                                Mu_sds = rep(10,9),
                                weight = pig.weights_mat,
                                use_lik = 1))
```

### Autoregressive (Order 1) covariance structure
```
Proc Mixed Data=Pigs Update;
    Class Week;
    Model Weight = Week / DDFM=KenwardRoger;
    Repeated Week / Subject=Pig Type=ARH(1) R RCorr RI;
    LSMeans Week;
Run;
```
$$\begin{align*}
  \mathbf{Y}_{i} &\overset{\text{iid}}{\sim} MVN(\boldsymbol{\mu},\boldsymbol{\Sigma}_{AR})
  \end{align*}
$$
The mean specification is the same. Now the variance covariance matrix $\boldsymbol{\Sigma}_{AR}$ is more structured. There are still 9 diagonal (variance) free parameters. The correlation matrix, which determines the off-diagonals along with the variance parameters, has the AR(1) structure. That is, neighboring outcomes (distance 1) within an individual have a correlation of $\rho$. Outcomes with a distance of 2 have a correlation of $\rho^2$. It keeps decaying until $Corr(Y_{i1},Y_{i9}) = \rho^8$. Therefore, the covariance matrix has 10 free parameters.

--------------------
- Top Page: http://rpubs.com/kaz_yos/
- Github: https://github.com/kaz-yos

```{r}
print(sessionInfo())
## Record execution time and multicore use
end_time <- Sys.time()
diff_time <- difftime(end_time, start_time, units = "auto")
cat("Started  ", as.character(start_time), "\n",
    "Finished ", as.character(end_time), "\n",
    "Time difference of ", diff_time, " ", attr(diff_time, "units"), "\n",
    "Used ", foreach::getDoParWorkers(), " cores\n",
    "Used ", foreach::getDoParName(), " as backend\n",
    sep = "")
```
